#Validation IHC CD163

```{r}
library(coxphf)
library(knitr)
library(openxlsx)
library(immunedeconv)
library(tidyr)
library(tibble)
library(survival)
library(dplyr)
library(survminer)
library(readxl)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(gridExtra)
library(scales)
```

```{r}
hpv_ind_xcell <- deconvolute(counts_HPV_Independent, "xcell")


hpv_ind_xcell_long <- hpv_ind_xcell %>%
  pivot_longer(
    cols = -cell_type, # All columns except `cell_type`
    names_to = "Sample", 
    values_to = "Fraction"
  )

analysis_data <- hpv_ind_xcell_long %>%
  left_join(TP53status_clinical_HPVind, by = c( "Sample"="id"))

```



```{r}
CD163 <- read_excel("Validation_CD163.xlsx", 
     col_types = c("text", "numeric", 
         "numeric", "numeric", "text"))

# Convert "NA" strings to actual NA in character columns
CD163 <- CD163 %>%
  mutate(across(where(is.character), ~ na_if(., "NA"))) %>%
  select(-Observations) %>%  # Remove the "Observations" column
  filter(rowSums(is.na(select(., -`Case_ID`))) < (ncol(.) - 1))  # Remove rows where all other columns are NA

# View cleaned dataset
print(CD163)



# Subset data for Macrophage M2
macrophage_m2_data <- analysis_data %>% 
  filter(cell_type == "Macrophage M2")


#We check the ones that are not in the other datset:
#Find which IDs are in CD163$Case_ID but not in macrophage_m2_data$Sample
setdiff(CD163$Case_ID, macrophage_m2_data$Sample)

#Find which IDs are in macrophage_m2_data$Sample but not in CD163$Case_ID
setdiff(macrophage_m2_data$Sample, CD163$Case_ID)

#2 cases


merged_data <- merge(CD163, macrophage_m2_data, 
                     by.x = "Case_ID", by.y = "Sample")

dim(merged_data)
#50 cases
```





```{r}
# Use anti_join to get cases from CD163 that don't have a match in macrophage_m2_data 
unmatched_cases <- CD163 %>%
  anti_join(macrophage_m2_data , by = c("Case_ID" = "Sample"))

# View the unmatched cases
print(unmatched_cases)

# Find cases in macrophage_m2_data that are NOT in CD163
unmatched_cases <- macrophage_m2_data %>%
  anti_join(CD163, by = c("Sample" = "Case_ID"))

# View the unmatched cases
print(unmatched_cases$Sample)
```









```{r}
# Ensure categorical variables are factors
merged_data <- merged_data %>%
  mutate(
    hpv_status = factor(hpv_status)
  )

```







#Differences between pdl1 positive and pdl1 negative tumors in CD163-positive cells
```{r}

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(ggpubr)
library(tidyr)

# Define the parameters of interest
cd163_params <- c(
  "Positive_cells_tumor_stroma",
  "Positive_cells_tumoral_area",
  "Positive_cells_perivascular_areas"
)

# Filter and reshape data to long format
plot_data <- merged_data %>%
  filter(pdl1_immune_cells_pos_category %in% c("Positive", "Negative")) %>%
  select(pdl1_immune_cells_pos_category, all_of(cd163_params)) %>%
  pivot_longer(cols = -pdl1_immune_cells_pos_category, 
               names_to = "CD163_Parameter", 
               values_to = "Value")

# Plot with facets
combined_plot <- ggplot(plot_data, aes(x = pdl1_immune_cells_pos_category, y = Value, fill = pdl1_immune_cells_pos_category)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, position = position_dodge(0.9)) +
  geom_jitter(aes(color = pdl1_immune_cells_pos_category),
              position = position_jitter(width = 0.1),
              size = 2, alpha = 0.6) +
  stat_compare_means(method = "wilcox.test",
                     label = "p.format",
                     label.x.npc = "center",
                     label.y.npc = "top",
                     size = 5) +
  facet_wrap(~CD163_Parameter, scales = "free_y") +
  scale_fill_manual(values = c("Negative" = "skyblue1", "Positive" = "skyblue4")) +
  scale_color_manual(values = c("Negative" = "dodgerblue3", "Positive" = "darkblue")) +
  theme_minimal() +
  labs(
    title = "CD163 Parameters by PD-L1 Immune Cell Status",
    x = "PD-L1 Status",
    y = "CD163-Positive Cell Count",
    fill = "PD-L1 Status",
    color = "PD-L1 Status"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

# Show plot
print(combined_plot)

# Save to file
ggsave("CD163_by_PDL1_Status_combined_plot.pdf", plot = combined_plot, width = 10, height = 7)
```



#REVIEW
```{r}
# ============================================================================
# CD163 IHC SURVIVAL ANALYSIS
# DSS + RFS
# maxstat + median + continuous
# ============================================================================

ihc_data <- merged_data %>%
  mutate(
    figo_ini_ava = relevel(
      factor(figo_ini_ava),
      ref = "1 (Advanced)"
    ),
    p53_ihc_status = relevel(
      factor(p53_ihc_status),
      ref = "Mutant"
    ),
    death_disease_60m = case_when(
      as.character(death_disease_60m) %in% c("0", "No")  ~ 0,
      as.character(death_disease_60m) %in% c("1", "Yes") ~ 1,
      TRUE ~ NA_real_
    ),
    recurrence_60m = case_when(
      as.character(recurrence_60m) %in% c("0", "No")  ~ 0,
      as.character(recurrence_60m) %in% c("1", "Yes") ~ 1,
      TRUE ~ NA_real_
    )
  )

# CD163 IHC markers / compartments
compartments <- c(
  "Positive_cells_tumoral_area",
  "Positive_cells_tumor_stroma",
  "Positive_cells_perivascular_areas"
)

# Covariates used in the CD163 multivariable models
covariates_full <- c(
  "figo_ini_ava",
  "p53_ihc_status",
  "pdl1_immune_cells_pos_category"
)
```



```{r}
# ============================================================================
# SHARED HELPERS
# ============================================================================

check_convergence <- function(model, maxit_val) {
  n_iter <- model$iter

  !is.null(n_iter) &&
    all(n_iter < maxit_val) &&
    is.finite(sum(model$coefficients))
}


find_maxstat_cutoff_ihc <- function(time_var, event_var, marker_var, data) {

  d <- data[
    !is.na(data[[marker_var]]) &
      !is.na(data[[time_var]]) &
      !is.na(data[[event_var]]),
  ]

  n <- nrow(d)

  if (n < 10) return(NA_real_)

  fmla <- as.formula(
    paste0(
      "survival::Surv(",
      time_var,
      ", ",
      event_var,
      ") ~ ",
      marker_var
    )
  )

  MAXSTAT <- tryCatch(
    maxstat::maxstat.test(
      fmla,
      data = d,
      smethod = "LogRank",
      pmethod = "Lau92",
      iscores = TRUE,
      minprop = 0.2,
      maxprop = 0.8
    ),
    error = function(e) NA
  )

  if (isTRUE(is.na(MAXSTAT))) return(NA_real_)

  MAXSTAT$estimate
}


add_threshold_groups_ihc <- function(
    df,
    time_var,
    event_var,
    marker_var
) {

  cutoff_maxstat <- tryCatch(
    find_maxstat_cutoff_ihc(
      time_var,
      event_var,
      marker_var,
      df
    ),
    error = function(e) NA_real_
  )

  cutoff_median <- median(
    df[[marker_var]],
    na.rm = TRUE
  )

  df %>%
    mutate(
      Group_maxstat = ifelse(
        !is.na(cutoff_maxstat) &
          .data[[marker_var]] > cutoff_maxstat,
        "High",
        "Low"
      ),

      Group_median = ifelse(
        .data[[marker_var]] > cutoff_median,
        "High",
        "Low"
      )
    )
}


make_cutoff_table_ihc <- function(data, markers, time_var, event_var) {
  bind_rows(lapply(markers, function(m) {
    d  <- data %>% filter(!is.na(.data[[m]]), !is.na(.data[[time_var]]), !is.na(.data[[event_var]]))
    ev <- as.numeric(as.character(d[[event_var]]))
    cuts <- c(maxstat = unname(find_maxstat_cutoff_ihc(time_var, event_var, m, d)),
              median  = median(d[[m]], na.rm = TRUE))
    bind_rows(lapply(names(cuts), function(tt) {
      high <- !is.na(cuts[[tt]]) & d[[m]] > cuts[[tt]]
      data.frame(marker = m, threshold_type = tt,
                 cutoff = signif(cuts[[tt]], 3),
                 `n (Low/High)`      = paste0(sum(!high), "/", sum(high)),
                 `Events (Low/High)` = paste0(sum(ev[!high] == 1), "/", sum(ev[high] == 1)),
                 check.names = FALSE)
    }))
  }))
}

add_cutoffs_ihc <- function(tab, cutoff_table) {
  tab <- tab %>%
    left_join(cutoff_table, by = c("marker", "threshold_type")) %>%
    relocate(cutoff, `n (Low/High)`, `Events (Low/High)`, .after = threshold_type)
  if ("Variable" %in% names(tab)) {
    keep <- grepl("^Enrichment", tab$Variable)
    tab[!keep, c("cutoff", "n (Low/High)", "Events (Low/High)")] <- NA
  }
  tab$converged <- ifelse(tab$converged, "Yes", "No")
  tab
}
```




```{r}
# ============================================================================
# DSS — UNIVARIATE
# ============================================================================

run_cox_DSS_ihc_group <- function(
    marker_label,
    marker_var,
    exposure_var,
    threshold_label,
    data,
    maxit_val = 100000
) {

  df0 <- add_threshold_groups_ihc(
    data,
    "time_death_surv_60m",
    "death_disease_60m",
    marker_var
  )

  df0[[exposure_var]] <- relevel(
    factor(df0[[exposure_var]]),
    ref = "Low"
  )

  df0 <- df0 %>%
    filter(
      !is.na(.data[[exposure_var]]),
      !is.na(death_disease_60m),
      !is.na(time_death_surv_60m)
    )

  fmla <- as.formula(
    paste0(
      "Surv(time_death_surv_60m, death_disease_60m) ~ ",
      exposure_var
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {
    return(
      data.frame(
        marker = marker_label,
        threshold_type = threshold_label,
        coef = NA,
        HR = NA,
        lower_095 = NA,
        upper_095 = NA,
        pval = NA,
        n = nrow(df0),
        n_events = sum(
          df0$death_disease_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE
      )
    )
  }

  s <- summary(model)

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    threshold_type = threshold_label,
    coef = s$coefficients,
    HR = exp(s$coefficients),
    lower_095 = s$ci.lower,
    upper_095 = s$ci.upper,
    pval = s$prob,
    n = nrow(df0),
    n_events = sum(
      df0$death_disease_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged
  )
}


run_cox_DSS_ihc_continuous <- function(
    marker_label,
    marker_var,
    data,
    maxit_val = 100000
) {

  df0 <- data %>%
    filter(
      !is.na(.data[[marker_var]]),
      !is.na(death_disease_60m),
      !is.na(time_death_surv_60m)
    )

  fmla <- as.formula(
    paste0(
      "Surv(time_death_surv_60m, death_disease_60m) ~ ",
      marker_var
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {
    return(
      data.frame(
        marker = marker_label,
        threshold_type = "continuous",
        coef = NA,
        HR = NA,
        lower_095 = NA,
        upper_095 = NA,
        pval = NA,
        n = nrow(df0),
        n_events = sum(
          df0$death_disease_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE
      )
    )
  }

  s <- summary(model)

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    threshold_type = "continuous",
    coef = s$coefficients,
    HR = exp(s$coefficients),
    lower_095 = s$ci.lower,
    upper_095 = s$ci.upper,
    pval = s$prob,
    n = nrow(df0),
    n_events = sum(
      df0$death_disease_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged
  )
}


univ_DSS_maxstat_ihc <- bind_rows(
  lapply(
    compartments,
    function(marker) {
      run_cox_DSS_ihc_group(
        marker,
        marker,
        "Group_maxstat",
        "maxstat",
        ihc_data
      )
    }
  )
)


univ_DSS_median_ihc <- bind_rows(
  lapply(
    compartments,
    function(marker) {
      run_cox_DSS_ihc_group(
        marker,
        marker,
        "Group_median",
        "median",
        ihc_data
      )
    }
  )
)


univ_DSS_continuous_ihc <- bind_rows(
  lapply(
    compartments,
    function(marker) {
      run_cox_DSS_ihc_continuous(
        marker,
        marker,
        ihc_data
      )
    }
  )
)


univ_DSS_ihc_all <- bind_rows(
  univ_DSS_maxstat_ihc,
  univ_DSS_median_ihc,
  univ_DSS_continuous_ihc
) %>%
  group_by(threshold_type) %>%
  mutate(
    p_adj_BH = p.adjust(
      pval,
      method = "BH"
    )
  ) %>%
  ungroup()


univ_DSS_ihc_formatted <- univ_DSS_ihc_all %>%
  mutate(
    HR_CI = sprintf(
      "%.3f (%.3f–%.3f)",
      HR,
      lower_095,
      upper_095
    ),
    pval = round(pval, 4),
    p_adj_BH = round(p_adj_BH, 4)
  ) %>%
  select(
    marker,
    threshold_type,
    HR_CI,
    pval,
    p_adj_BH,
    n,
    n_events,
    n_iter,
    converged
  ) %>%
  arrange(
    marker,
    threshold_type
  )

cutoffs_DSS_cd163 <- make_cutoff_table_ihc(ihc_data, compartments,
                                           "time_death_surv_60m", "death_disease_60m")
cutoffs_DSS_cd163

univ_DSS_ihc_formatted <- add_cutoffs_ihc(univ_DSS_ihc_formatted, cutoffs_DSS_cd163)
kable(univ_DSS_ihc_formatted, caption = "Univariate Firth Cox regression for DSS: CD163 IHC")
```


```{r}
# ============================================================================
# DSS — SELECT MARKERS FOR MULTIVARIATE ANALYSIS
# Select the UNION of markers with BH-adjusted p < 0.10
# in ANY of the three univariate approaches
# ============================================================================

selected_DSS_ihc <- univ_DSS_ihc_all %>%
  filter(
    !is.na(p_adj_BH),
    p_adj_BH < 0.10
  ) %>%
  pull(marker) %>%
  unique()

cat(
  "CD163 markers entering DSS multivariable models:\n"
)

print(selected_DSS_ihc)
```

```{r}
# ============================================================================
# DSS — MULTIVARIATE MAXSTAT / MEDIAN
# adjusted for FIGO + p53
# ============================================================================

perform_cox_DSS_ihc_multiv <- function(
    marker_label,
    marker_var,
    exposure_var,
    data,
    covariates,
    maxit_val = 100000
) {

  df0 <- add_threshold_groups_ihc(
    data,
    "time_death_surv_60m",
    "death_disease_60m",
    marker_var
  )

  df0[[exposure_var]] <- relevel(
    factor(df0[[exposure_var]]),
    ref = "Low"
  )

  df0 <- df0 %>%
    filter(
      !is.na(.data[[exposure_var]]),
      !is.na(death_disease_60m),
      !is.na(time_death_surv_60m),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_death_surv_60m, death_disease_60m) ~ ",
      paste(
        c(exposure_var, covariates),
        collapse = " + "
      )
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {
    return(
      data.frame(
        marker = marker_label,
        variable = NA_character_,
        coef = NA_real_,
        HR = NA_real_,
        lower_095 = NA_real_,
        upper_095 = NA_real_,
        pval = NA_real_,
        n = nrow(df0),
        n_events = sum(
          df0$death_disease_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE,
        stringsAsFactors = FALSE
      )
    )
  }

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    variable = names(model$coefficients),
    coef = as.numeric(model$coefficients),
    HR = exp(as.numeric(model$coefficients)),
    lower_095 = as.numeric(model$ci.lower),
    upper_095 = as.numeric(model$ci.upper),
    pval = as.numeric(model$prob),
    n = nrow(df0),
    n_events = sum(
      df0$death_disease_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged,
    stringsAsFactors = FALSE
  )
}


multiv_DSS_maxstat_ihc <- if (
  length(selected_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_DSS_ihc,
      function(marker) {
        perform_cox_DSS_ihc_multiv(
          marker,
          marker,
          "Group_maxstat",
          ihc_data,
          covariates_full
        )
      }
    )
  ) %>%
    mutate(
      threshold_type = "maxstat",
      .after = marker
    )

} else {
  data.frame()
}


multiv_DSS_median_ihc <- if (
  length(selected_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_DSS_ihc,
      function(marker) {
        perform_cox_DSS_ihc_multiv(
          marker,
          marker,
          "Group_median",
          ihc_data,
          covariates_full
        )
      }
    )
  ) %>%
    mutate(
      threshold_type = "median",
      .after = marker
    )

} else {
  data.frame()
}
```






```{r}
# ============================================================================
# DSS — MULTIVARIATE CONTINUOUS
# ============================================================================

perform_cox_DSS_ihc_multiv_continuous <- function(
    marker_label,
    marker_var,
    data,
    covariates,
    maxit_val = 100000
) {

  df0 <- data %>%
    filter(
      !is.na(.data[[marker_var]]),
      !is.na(death_disease_60m),
      !is.na(time_death_surv_60m),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_death_surv_60m, death_disease_60m) ~ ",
      marker_var,
      " + ",
      paste(
        covariates,
        collapse = " + "
      )
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {
    return(
      data.frame(
        marker = marker_label,
        variable = NA_character_,
        coef = NA_real_,
        HR = NA_real_,
        lower_095 = NA_real_,
        upper_095 = NA_real_,
        pval = NA_real_,
        n = nrow(df0),
        n_events = sum(
          df0$death_disease_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE,
        stringsAsFactors = FALSE
      )
    )
  }

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    variable = names(model$coefficients),
    coef = as.numeric(model$coefficients),
    HR = exp(as.numeric(model$coefficients)),
    lower_095 = as.numeric(model$ci.lower),
    upper_095 = as.numeric(model$ci.upper),
    pval = as.numeric(model$prob),
    n = nrow(df0),
    n_events = sum(
      df0$death_disease_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged,
    stringsAsFactors = FALSE
  )
}


multiv_DSS_continuous_ihc <- if (
  length(selected_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_DSS_ihc,
      function(marker) {
        perform_cox_DSS_ihc_multiv_continuous(
          marker,
          marker,
          ihc_data,
          covariates_full
        )
      }
    )
  ) %>%
    mutate(
      threshold_type = "continuous",
      .after = marker
    )

} else {
  data.frame()
}


# BH correction for the marker term only
if (nrow(multiv_DSS_continuous_ihc) > 0) {

  multiv_DSS_continuous_ihc <- multiv_DSS_continuous_ihc %>%
    mutate(
      p_adj_BH = NA_real_
    )

  marker_rows <-
    multiv_DSS_continuous_ihc$variable ==
    multiv_DSS_continuous_ihc$marker

  multiv_DSS_continuous_ihc$p_adj_BH[marker_rows] <-
    p.adjust(
      multiv_DSS_continuous_ihc$pval[marker_rows],
      method = "BH"
    )
}
```


```{r}
# ============================================================================
# COMBINE DSS MULTIVARIATE RESULTS
# ============================================================================

multiv_DSS_ihc_all <- bind_rows(
  multiv_DSS_maxstat_ihc,
  multiv_DSS_median_ihc,
  multiv_DSS_continuous_ihc
)


# BH correction separately within each threshold,
# and only for the marker/exposure term
multiv_DSS_ihc_all <- multiv_DSS_ihc_all %>%
  group_by(threshold_type) %>%
  mutate(
    p_adj_BH = ifelse(
      grepl("^Group_", variable) |
        variable == marker,

      p.adjust(
        pval[
          grepl("^Group_", variable) |
            variable == marker
        ],
        method = "BH"
      )[match(
        pval,
        pval[
          grepl("^Group_", variable) |
            variable == marker
        ]
      )],

      NA_real_
    )
  ) %>%
  ungroup()


format_table_DSS_ihc <- function(df) {

  df %>%
    mutate(
      variable_label = case_when(
        grepl("^Group_", variable) ~
          "Enrichment (High vs Low)",

        variable == marker ~
          "Continuous (per unit)",

        variable == "figo_ini_ava0 (Initial)" ~
          "FIGO (Initial)",

        variable == "p53_ihc_statusWt" ~
          "p53 (Normal)",

        TRUE ~ variable
      ),

      HR_CI = sprintf(
        "%.4f (%.4f–%.4f)",
        HR,
        lower_095,
        upper_095
      ),

      coef = round(coef, 4),
      pval = round(pval, 4),
      p_adj_BH = round(p_adj_BH, 4)
    ) %>%
    select(
      marker,
      threshold_type,
      Variable = variable_label,
      Coef = coef,
      `HR (95% CI)` = HR_CI,
      `p-value` = pval,
      `p (BH-adj)` = p_adj_BH,
      n,
      n_events,
      n_iter,
      converged
    )
}


multiv_DSS_ihc_formatted <-
  format_table_DSS_ihc(
    multiv_DSS_ihc_all
  ) %>%
  arrange(
    marker,
    threshold_type
  )


multiv_DSS_ihc_formatted <- add_cutoffs_ihc(multiv_DSS_ihc_formatted, cutoffs_DSS_cd163)

kable(
  multiv_DSS_ihc_formatted,
  caption = "Multivariate Firth Cox regression for DSS: CD163 IHC"
)
```






```{r}
# ============================================================================
# DSS — PROPORTIONAL HAZARDS ASSUMPTION
# Schoenfeld residuals
# ============================================================================

check_ph_DSS_ihc <- function(
    marker_label,
    marker_var,
    data,
    exposure_var,
    covariates
) {

  df0 <- add_threshold_groups_ihc(
    data,
    "time_death_surv_60m",
    "death_disease_60m",
    marker_var
  )

  df0[[exposure_var]] <- relevel(
    factor(df0[[exposure_var]]),
    ref = "Low"
  )

  d <- df0 %>%
    mutate(
      death_disease_60m =
        as.numeric(
          as.character(death_disease_60m)
        )
    ) %>%
    filter(
      !is.na(.data[[exposure_var]]),
      !is.na(time_death_surv_60m),
      !is.na(death_disease_60m),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_death_surv_60m, death_disease_60m) ~ ",
      paste(
        c(exposure_var, covariates),
        collapse = " + "
      )
    )
  )

  fit <- tryCatch(
    coxph(
      fmla,
      data = d
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$death_disease_60m == 1,
          na.rm = TRUE
        ),
        note = "coxph_fit_failed"
      )
    )
  }

  zph <- tryCatch(
    cox.zph(fit),
    error = function(e) NULL
  )

  if (is.null(zph)) {
    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$death_disease_60m == 1,
          na.rm = TRUE
        ),
        note = "zph_failed"
      )
    )
  }

  tab <- as.data.frame(
    zph$table
  )

  tab$term <- rownames(tab)

  global_p <- tab$p[
    tab$term == "GLOBAL"
  ]

  data.frame(
    marker = marker_label,
    term = tab$term,
    chisq = round(tab$chisq, 3),
    df = tab$df,
    p = round(tab$p, 4),
    global_p = round(global_p, 4),
    n = nrow(d),
    n_events = sum(
      d$death_disease_60m == 1,
      na.rm = TRUE
    ),
    note = NA_character_
  )
}


ph_DSS_maxstat_ihc <- if (
  length(selected_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_DSS_ihc,
      function(m) {
        check_ph_DSS_ihc(
          m,
          m,
          ihc_data,
          "Group_maxstat",
          covariates_full
        )
      }
    )
  ) %>%
    mutate(
      threshold_type = "maxstat",
      .after = marker
    )

} else {
  data.frame()
}


ph_DSS_median_ihc <- if (
  length(selected_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_DSS_ihc,
      function(m) {
        check_ph_DSS_ihc(
          m,
          m,
          ihc_data,
          "Group_median",
          covariates_full
        )
      }
    )
  ) %>%
    mutate(
      threshold_type = "median",
      .after = marker
    )

} else {
  data.frame()
}


# Continuous PH function
check_ph_DSS_ihc_continuous <- function(
    marker_label,
    marker_var,
    data,
    covariates
) {

  d <- data %>%
    filter(
      !is.na(.data[[marker_var]]),
      !is.na(time_death_surv_60m),
      !is.na(death_disease_60m),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    mutate(
      death_disease_60m =
        as.numeric(
          as.character(death_disease_60m)
        )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_death_surv_60m, death_disease_60m) ~ ",
      marker_var,
      " + ",
      paste(
        covariates,
        collapse = " + "
      )
    )
  )

  fit <- tryCatch(
    coxph(
      fmla,
      data = d
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$death_disease_60m == 1,
          na.rm = TRUE
        ),
        note = "coxph_fit_failed"
      )
    )
  }

  zph <- tryCatch(
    cox.zph(fit),
    error = function(e) NULL
  )

  if (is.null(zph)) {
    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$death_disease_60m == 1,
          na.rm = TRUE
        ),
        note = "zph_failed"
      )
    )
  }

  tab <- as.data.frame(
    zph$table
  )

  tab$term <- rownames(tab)

  global_p <- tab$p[
    tab$term == "GLOBAL"
  ]

  data.frame(
    marker = marker_label,
    term = tab$term,
    chisq = round(tab$chisq, 3),
    df = tab$df,
    p = round(tab$p, 4),
    global_p = round(global_p, 4),
    n = nrow(d),
    n_events = sum(
      d$death_disease_60m == 1,
      na.rm = TRUE
    ),
    note = NA_character_
  )
}


ph_DSS_continuous_ihc <- if (
  length(selected_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_DSS_ihc,
      function(m) {
        check_ph_DSS_ihc_continuous(
          m,
          m,
          ihc_data,
          covariates_full
        )
      }
    )
  ) %>%
    mutate(
      threshold_type = "continuous",
      .after = marker
    )

} else {
  data.frame()
}


ph_DSS_ihc_all <- bind_rows(
  ph_DSS_maxstat_ihc,
  ph_DSS_median_ihc,
  ph_DSS_continuous_ihc
) %>%
  arrange(
    marker,
    threshold_type,
    term
  )

kable(
  ph_DSS_ihc_all,
  caption =
    "Proportional hazards assumption for multivariable DSS models: CD163 IHC"
)
```

```{r}

# ============================================================================
# DSS — SELECT MULTIVARIABLE MODELS FOR BOOTSTRAP
# BH-adjusted p < 0.05
# ============================================================================

sig_maxstat_DSS_ihc <- multiv_DSS_ihc_all %>%
  filter(
    threshold_type == "maxstat",
    grepl("^Group_maxstat", variable),
    !is.na(p_adj_BH),
    p_adj_BH < 0.05
  ) %>%
  pull(marker) %>%
  unique()


sig_median_DSS_ihc <- multiv_DSS_ihc_all %>%
  filter(
    threshold_type == "median",
    grepl("^Group_median", variable),
    !is.na(p_adj_BH),
    p_adj_BH < 0.05
  ) %>%
  pull(marker) %>%
  unique()


sig_continuous_DSS_ihc <- multiv_DSS_ihc_all %>%
  filter(
    threshold_type == "continuous",
    variable == marker,
    !is.na(p_adj_BH),
    p_adj_BH < 0.05
  ) %>%
  pull(marker) %>%
  unique()


cat("DSS maxstat models for bootstrap:\n")
print(sig_maxstat_DSS_ihc)

cat("\nDSS median models for bootstrap:\n")
print(sig_median_DSS_ihc)

cat("\nDSS continuous models for bootstrap:\n")
print(sig_continuous_DSS_ihc)
```


```{r}
# ============================================================================
# DSS — BOOTSTRAP OPTIMISM CORRECTION
# ============================================================================

get_cutoff_DSS_ihc <- function(
    threshold_type,
    marker_var,
    data
) {

  if (threshold_type == "maxstat") {

    find_maxstat_cutoff_ihc(
      "time_death_surv_60m",
      "death_disease_60m",
      marker_var,
      data
    )

  } else {

    median(
      data[[marker_var]],
      na.rm = TRUE
    )
  }
}


# ============================================================================
# DSS — BOOTSTRAP OPTIMISM CORRECTION
# ============================================================================

bootstrap_DSS_ihc <- function(
    marker_val,
    threshold_type,
    data,
    covariates,
    B = 1000,
    maxit_boot = 500,
    maxstep_boot = 0.5,
    seed = 1
) {

  t0 <- Sys.time()

  # --------------------------------------------------------------------------
  # Original complete-case dataset
  # --------------------------------------------------------------------------

  df0 <- data %>%
    filter(
      !is.na(.data[[marker_val]]),
      !is.na(time_death_surv_60m),
      !is.na(death_disease_60m),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    droplevels()


  # --------------------------------------------------------------------------
  # Apparent cutoff
  # --------------------------------------------------------------------------

  cutoff0 <- if (
    threshold_type == "continuous"
  ) {

    NA_real_

  } else {

    get_cutoff_DSS_ihc(
      threshold_type,
      marker_val,
      df0
    )
  }


  # --------------------------------------------------------------------------
  # Build exposure
  # --------------------------------------------------------------------------

  if (threshold_type != "continuous") {

    if (is.na(cutoff0)) {

      return(
        data.frame(
          marker = marker_val,
          threshold_type = threshold_type,
          HR_apparent = NA,
          HR_corrected = NA,
          HR_corrected_lower = NA,
          HR_corrected_upper = NA,
          optimism = NA,
          B = B,
          n_valid_boot = 0,
          pct_valid = 0,
          elapsed_sec = NA,
          error = "cutoff_not_found"
        )
      )
    }

    df0$Group <- factor(
      ifelse(
        df0[[marker_val]] > cutoff0,
        "High",
        "Low"
      ),
      levels = c("Low", "High")
    )

    if (
      nlevels(
        droplevels(df0$Group)
      ) < 2
    ) {

      return(
        data.frame(
          marker = marker_val,
          threshold_type = threshold_type,
          HR_apparent = NA,
          HR_corrected = NA,
          HR_corrected_lower = NA,
          HR_corrected_upper = NA,
          optimism = NA,
          B = B,
          n_valid_boot = 0,
          pct_valid = 0,
          elapsed_sec = NA,
          error = "single_group"
        )
      )
    }

    # Variable used in formula
    exposure_var <- "Group"

    # IMPORTANT:
    # coxphf() names the coefficient "GroupHigh"
    coefficient_name <- "GroupHigh"

  } else {

    exposure_var <- marker_val

    # Continuous variable has same name as coefficient
    coefficient_name <- marker_val
  }


  # --------------------------------------------------------------------------
  # Formula
  # --------------------------------------------------------------------------

  fmla_boot <- as.formula(
    paste0(
      "Surv(time_death_surv_60m, death_disease_60m) ~ ",
      exposure_var,
      " + ",
      paste(
        covariates,
        collapse = " + "
      )
    )
  )


  # --------------------------------------------------------------------------
  # Apparent model
  # --------------------------------------------------------------------------

  fit_apparent <- tryCatch(
    coxphf(
      fmla_boot,
      data = df0,
      maxstep = 0.0001,
      maxit = 100000,
      pl = TRUE
    ),
    error = function(e) NULL
  )


  # --------------------------------------------------------------------------
  # Check fitted model
  # --------------------------------------------------------------------------

  if (is.null(fit_apparent)) {

    return(
      data.frame(
        marker = marker_val,
        threshold_type = threshold_type,
        HR_apparent = NA,
        HR_corrected = NA,
        HR_corrected_lower = NA,
        HR_corrected_upper = NA,
        optimism = NA,
        B = B,
        n_valid_boot = 0,
        pct_valid = 0,
        elapsed_sec = NA,
        error = "apparent_fit_failed"
      )
    )
  }


  # --------------------------------------------------------------------------
  # Check coefficient
  # --------------------------------------------------------------------------

  if (
    !(coefficient_name %in% names(fit_apparent$coefficients))
  ) {

    return(
      data.frame(
        marker = marker_val,
        threshold_type = threshold_type,
        HR_apparent = NA,
        HR_corrected = NA,
        HR_corrected_lower = NA,
        HR_corrected_upper = NA,
        optimism = NA,
        B = B,
        n_valid_boot = 0,
        pct_valid = 0,
        elapsed_sec = NA,
        error = paste0(
          "coefficient_not_found: ",
          coefficient_name
        )
      )
    )
  }


  beta_apparent <- unname(
    fit_apparent$coefficients[
      coefficient_name
    ]
  )


  # --------------------------------------------------------------------------
  # Bootstrap
  # --------------------------------------------------------------------------

  set.seed(seed)

  n <- nrow(df0)

  optimism_vec <- rep(
    NA_real_,
    B
  )


  for (b in seq_len(B)) {

    idx <- sample(
      seq_len(n),
      n,
      replace = TRUE
    )

    dboot <- df0[idx, ]


    # ------------------------------------------------------------------------
    # Recalculate cutoff in each bootstrap sample
    # ------------------------------------------------------------------------

    if (threshold_type != "continuous") {

      cutoff_b <- get_cutoff_DSS_ihc(
        threshold_type,
        marker_val,
        dboot
      )

      if (is.na(cutoff_b)) next


      dboot$Group <- factor(
        ifelse(
          dboot[[marker_val]] > cutoff_b,
          "High",
          "Low"
        ),
        levels = c("Low", "High")
      )


      if (
        nlevels(
          droplevels(dboot$Group)
        ) < 2
      ) next
      
      ev_tab_boot <- table(dboot$Group, dboot$death_disease_60m)
      if (ncol(ev_tab_boot) < 2 || any(ev_tab_boot[, 2] == 0)) next
    }


    # ------------------------------------------------------------------------
    # Fit bootstrap model
    # ------------------------------------------------------------------------

    fit_boot <- tryCatch(
      coxphf(
        fmla_boot,
        data = dboot,
        maxstep = maxstep_boot,
        maxit = maxit_boot,
        pl = FALSE
      ),
      error = function(e) NULL
    )


    if (is.null(fit_boot)) next


    if (
      !(coefficient_name %in% names(fit_boot$coefficients))
    ) next


    beta_boot_apparent <- unname(
      fit_boot$coefficients[
        coefficient_name
      ]
    )


    # ------------------------------------------------------------------------
    # Test bootstrap model on original data
    # ------------------------------------------------------------------------

    if (threshold_type != "continuous") {

      df_test <- df0

      # IMPORTANT:
      # Use the bootstrap-derived cutoff on the original data
      df_test$Group <- factor(
        ifelse(
          df_test[[marker_val]] > cutoff_b,
          "High",
          "Low"
        ),
        levels = c("Low", "High")
      )

      if (
        nlevels(
          droplevels(df_test$Group)
        ) < 2
      ) next      
      
      ev_tab_test <- table(df_test$Group, df_test$death_disease_60m)
      if (ncol(ev_tab_test) < 2 || any(ev_tab_test[, 2] == 0)) next


    } else {

      df_test <- df0

    }


    fit_test <- tryCatch(
      coxphf(
        fmla_boot,
        data = df_test,
        maxstep = maxstep_boot,
        maxit = maxit_boot,
        pl = FALSE
      ),
      error = function(e) NULL
    )


    if (is.null(fit_test)) next


    if (
      !(coefficient_name %in% names(fit_test$coefficients))
    ) next


    beta_boot_test <- unname(
      fit_test$coefficients[
        coefficient_name
      ]
    )


    # ------------------------------------------------------------------------
    # Optimism
    # ------------------------------------------------------------------------

    optimism_vec[b] <-
      beta_boot_apparent -
      beta_boot_test
  }


  # --------------------------------------------------------------------------
  # Valid bootstrap samples
  # --------------------------------------------------------------------------

  valid_optimism <-
    optimism_vec[
      !is.na(optimism_vec)
    ]

  n_valid <-
    length(valid_optimism)

  elapsed <-
    as.numeric(
      difftime(
        Sys.time(),
        t0,
        units = "secs"
      )
    )


  # --------------------------------------------------------------------------
  # Require sufficient valid bootstrap samples
  # --------------------------------------------------------------------------

  if (
    n_valid < max(
      10,
      0.2 * B
    )
  ) {

    return(
      data.frame(
        marker = marker_val,
        threshold_type = threshold_type,
        HR_apparent = exp(beta_apparent),
        HR_corrected = NA,
        HR_corrected_lower = NA,
        HR_corrected_upper = NA,
        optimism = NA,
        B = B,
        n_valid_boot = n_valid,
        pct_valid = round(
          100 * n_valid / B,
          1
        ),
        elapsed_sec = round(
          elapsed,
          1
        ),
        error = "too_few_valid_bootstraps"
      )
    )
  }


  # --------------------------------------------------------------------------
  # Mean optimism
  # --------------------------------------------------------------------------

  mean_optimism <-
    mean(
      valid_optimism
    )


  beta_corrected <-
    beta_apparent -
    mean_optimism


  corrected_dist <-
    beta_apparent -
    valid_optimism


  ci <-
    quantile(
      corrected_dist,
      probs = c(
        0.025,
        0.975
      ),
      na.rm = TRUE
    )


  # --------------------------------------------------------------------------
  # Final result
  # --------------------------------------------------------------------------

  data.frame(
    marker = marker_val,
    threshold_type = threshold_type,
    HR_apparent = exp(beta_apparent),
    HR_corrected = exp(beta_corrected),
    HR_corrected_lower = exp(ci[1]),
    HR_corrected_upper = exp(ci[2]),
    optimism = mean_optimism,
    B = B,
    n_valid_boot = n_valid,
    pct_valid = round(
      100 * n_valid / B,
      1
    ),
    elapsed_sec = round(
      elapsed,
      1
    ),
    error = NA_character_
  )
}
```


```{r}
test_median <- bootstrap_DSS_ihc(
  marker_val = "Positive_cells_tumoral_area",
  threshold_type = "median",
  data = ihc_data,
  covariates = covariates_full,
  B = 20,
  seed = 123
)

test_median
```



```{r}
test_maxstat <- bootstrap_DSS_ihc(
  marker_val = "Positive_cells_tumoral_area",
  threshold_type = "maxstat",
  data = ihc_data,
  covariates = covariates_full,
  B = 20,
  seed = 123
)

test_maxstat
```
```{r}
test_maxstat_tumoral_area <- bootstrap_DSS_ihc(
  marker_val = "Positive_cells_tumoral_area",
  threshold_type = "maxstat",
  data = ihc_data,
  covariates = covariates_full,
  B = 20,
  seed = 123
)

test_maxstat_tumoral_area
```

```{r}
boot_DSS_ihc_maxstat <- if (
  length(sig_maxstat_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      sig_maxstat_DSS_ihc,
      bootstrap_DSS_ihc,
      threshold_type = "maxstat",
      data = ihc_data,
      covariates = covariates_full,
      B = 1000
    )
  )

} else {
  data.frame()
}


boot_DSS_ihc_median <- if (
  length(sig_median_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      sig_median_DSS_ihc,
      bootstrap_DSS_ihc,
      threshold_type = "median",
      data = ihc_data,
      covariates = covariates_full,
      B = 1000
    )
  )

} else {
  data.frame()
}


boot_DSS_ihc_continuous <- if (
  length(sig_continuous_DSS_ihc) > 0
) {

  bind_rows(
    lapply(
      sig_continuous_DSS_ihc,
      bootstrap_DSS_ihc,
      threshold_type = "continuous",
      data = ihc_data,
      covariates = covariates_full,
      B = 1000
    )
  )

} else {
  data.frame()
}


boot_DSS_ihc_all <- bind_rows(
  boot_DSS_ihc_maxstat,
  boot_DSS_ihc_median,
  boot_DSS_ihc_continuous
) %>%
  arrange(
    marker,
    threshold_type
  )

boot_DSS_ihc_all
```



#Visualization
```{r}

build_marker_order <- function(markers) {
  markers <- unique(markers)
  is_score <- grepl("score", markers, ignore.case = TRUE)
  ordered <- c(sort(markers[is_score]), sort(markers[!is_score]))
  rev(ordered)
}
```


```{r}
# ============================================================================
# FOREST PLOT -- DSS (CD163 IHC), maxstat vs. median, bootstrap-corrected HRs
# ============================================================================

library(ggplot2)

make_optimism_forest_ihc <- function(boot_df, multiv_df, title) {

  sig_df <- multiv_df %>%
    filter(Variable == "Enrichment (High vs Low)") %>%
    transmute(marker, threshold_type,
              significant = !is.na(`p (BH-adj)`) & `p (BH-adj)` < 0.05)

  df <- boot_df %>%
    filter(!is.na(HR_corrected), threshold_type %in% c("maxstat", "median")) %>%
    left_join(sig_df, by = c("marker", "threshold_type")) %>%
    mutate(
      significant = ifelse(is.na(significant), FALSE, significant),
      threshold_type = factor(threshold_type, levels = c("maxstat", "median"),
                               labels = c("Maxstat", "Median"))
    )

  marker_order <- build_marker_order(df$marker)
  df <- df %>% mutate(marker = factor(marker, levels = marker_order))

  ggplot(df, aes(x = marker, y = HR_corrected, color = threshold_type,
                  shape = significant)) +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    geom_errorbar(aes(ymin = HR_corrected_lower, ymax = HR_corrected_upper),
                  position = position_dodge(width = 0.5), width = 0.2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
    coord_flip() +
    scale_y_log10() +
    scale_color_manual(values = c("Maxstat" = "#D55E00", "Median" = "#0072B2"),
                        name = "Threshold") +
    scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                        labels = c("TRUE" = "p.adj < 0.05", "FALSE" = "n.s."),
                        name = "Significance\n(apparent model)") +
    labs(title = title, x = "", y = "Bootstrap-corrected Hazard Ratio (log scale)") +
    theme_minimal()
}

forest_DSS_ihc_corrected <- make_optimism_forest_ihc(
  boot_DSS_ihc_all, multiv_DSS_ihc_formatted,
  "DSS \u2014 CD163 IHC, bootstrap-corrected HRs, maxstat vs. median"
)
forest_DSS_ihc_corrected
```

```{r}
# ============================================================================
# FOREST PLOT -- DSS (CD163 IHC), maxstat vs. median, NON-BOOTSTRAP HRs
# ============================================================================

make_forest_ihc <- function(multiv_df, multiv_raw, title) {

  sig_df <- multiv_df %>%
    filter(Variable == "Enrichment (High vs Low)") %>%
    transmute(marker, threshold_type,
              significant = !is.na(`p (BH-adj)`) & `p (BH-adj)` < 0.05)

  df <- multiv_raw %>%
    filter(variable %in% c("Group_maxstatHigh", "Group_medianHigh")) %>%
    left_join(sig_df, by = c("marker", "threshold_type")) %>%
    mutate(
      significant = ifelse(is.na(significant), FALSE, significant),
      threshold_type = factor(threshold_type, levels = c("maxstat", "median"),
                               labels = c("Maxstat", "Median"))
    ) %>%
    filter(!is.na(HR), !is.na(lower_095), !is.na(upper_095))

  marker_order <- build_marker_order(df$marker)
  df <- df %>% mutate(marker = factor(marker, levels = marker_order))

  ggplot(df, aes(x = marker, y = HR, color = threshold_type, shape = significant)) +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    geom_errorbar(aes(ymin = lower_095, ymax = upper_095),
                  position = position_dodge(width = 0.5), width = 0.2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
    coord_flip() +
    scale_y_log10() +
    scale_color_manual(values = c("Maxstat" = "#D55E00", "Median" = "#0072B2"), name = "Threshold") +
    scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                        labels = c("TRUE" = "p.adj < 0.05", "FALSE" = "n.s."), name = "Significance") +
    labs(title = title, x = "", y = "Hazard Ratio (log scale)") +
    theme_minimal()
}

forest_DSS_ihc_unadjusted <- make_forest_ihc(
  multiv_df  = multiv_DSS_ihc_formatted,
  multiv_raw = multiv_DSS_ihc_all,
  title = "DSS \u2014 CD163 IHC, multivariable HRs, maxstat vs. median"
)

forest_DSS_ihc_unadjusted
```



```{r}
# ============================================================================
# KAPLAN-MEIER -- DSS by CD163 enrichment (maxstat and median)
# ============================================================================

plot_km_ihc <- function(data, marker_var, time_var, event_var, group_var, title) {

  df0 <- add_threshold_groups_ihc(data, time_var, event_var, marker_var)
  df0[[event_var]] <- as.numeric(as.character(df0[[event_var]]))
  df0[[group_var]] <- droplevels(factor(df0[[group_var]], levels = c("Low", "High")))
  df0 <- df0 %>%
    filter(!is.na(.data[[group_var]]), !is.na(.data[[time_var]]), !is.na(.data[[event_var]]))

  grp_counts <- table(df0[[group_var]])
  if (any(grp_counts == 0) || length(grp_counts) < 2) {
    stop(sprintf(
      "Cannot build KM plot for '%s' / %s: group counts are %s.",
      marker_var, group_var, paste(names(grp_counts), grp_counts, sep = "=", collapse = ", ")
    ))
  }

  fmla <- as.formula(paste0("Surv(", time_var, ", ", event_var, ") ~ ", group_var))
  fit <- survminer::surv_fit(fmla, data = df0)

  ggsurvplot(
    fit, data = df0, pval = TRUE, risk.table = TRUE,
    legend.title = paste(marker_var, "enrichment"), legend.labs = c("Low", "High"),
    title = title, palette = c("gray50", "black"),
    xlab = "Time (months)", ylab = "Survival probability"
  )
}

km_dss_ihc_maxstat <- lapply(sig_maxstat_DSS_ihc, function(m) {
  plot_km_ihc(ihc_data, m, "time_death_surv_60m", "death_disease_60m", "Group_maxstat",
              paste0("DSS by ", m, " enrichment (maxstat)"))
})

km_dss_ihc_median <- lapply(sig_median_DSS_ihc, function(m) {
  plot_km_ihc(ihc_data, m, "time_death_surv_60m", "death_disease_60m", "Group_median",
              paste0("DSS by ", m, " enrichment (median)"))
})

km_dss_ihc_maxstat
km_dss_ihc_median
```









#Recurrence
```{r}
# ============================================================================
# RFS — CD163 IHC SURVIVAL ANALYSIS
# maxstat + median + continuous
# univariate -> multivariate -> PH test -> bootstrap
# ============================================================================


# ============================================================================
# 1. RFS — UNIVARIATE
# ============================================================================

run_cox_RFS_ihc_group <- function(
    marker_label,
    marker_var,
    exposure_var,
    threshold_label,
    data,
    maxit_val = 100000
) {

  df0 <- add_threshold_groups_ihc(
    data,
    "time_recurrence_surv_60m",
    "recurrence_60m",
    marker_var
  )

  df0[[exposure_var]] <- relevel(
    factor(df0[[exposure_var]]),
    ref = "Low"
  )

  df0 <- df0 %>%
    filter(
      !is.na(.data[[exposure_var]]),
      !is.na(recurrence_60m),
      !is.na(time_recurrence_surv_60m)
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_recurrence_surv_60m, recurrence_60m) ~ ",
      exposure_var
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {

    return(
      data.frame(
        marker = marker_label,
        threshold_type = threshold_label,
        coef = NA,
        HR = NA,
        lower_095 = NA,
        upper_095 = NA,
        pval = NA,
        n = nrow(df0),
        n_events = sum(
          df0$recurrence_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE
      )
    )
  }

  s <- summary(model)

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    threshold_type = threshold_label,
    coef = s$coefficients,
    HR = exp(s$coefficients),
    lower_095 = s$ci.lower,
    upper_095 = s$ci.upper,
    pval = s$prob,
    n = nrow(df0),
    n_events = sum(
      df0$recurrence_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged
  )
}


# ----------------------------------------------------------------------------
# RFS — UNIVARIATE CONTINUOUS
# ----------------------------------------------------------------------------

run_cox_RFS_ihc_continuous <- function(
    marker_label,
    marker_var,
    data,
    maxit_val = 100000
) {

  df0 <- data %>%
    filter(
      !is.na(.data[[marker_var]]),
      !is.na(recurrence_60m),
      !is.na(time_recurrence_surv_60m)
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_recurrence_surv_60m, recurrence_60m) ~ ",
      marker_var
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {

    return(
      data.frame(
        marker = marker_label,
        threshold_type = "continuous",
        coef = NA,
        HR = NA,
        lower_095 = NA,
        upper_095 = NA,
        pval = NA,
        n = nrow(df0),
        n_events = sum(
          df0$recurrence_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE
      )
    )
  }

  s <- summary(model)

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    threshold_type = "continuous",
    coef = s$coefficients,
    HR = exp(s$coefficients),
    lower_095 = s$ci.lower,
    upper_095 = s$ci.upper,
    pval = s$prob,
    n = nrow(df0),
    n_events = sum(
      df0$recurrence_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged
  )
}


# ----------------------------------------------------------------------------
# Run all three RFS univariate analyses
# ----------------------------------------------------------------------------

univ_RFS_maxstat_ihc <- bind_rows(
  lapply(
    compartments,
    function(marker) {

      run_cox_RFS_ihc_group(
        marker,
        marker,
        "Group_maxstat",
        "maxstat",
        ihc_data
      )

    }
  )
)


univ_RFS_median_ihc <- bind_rows(
  lapply(
    compartments,
    function(marker) {

      run_cox_RFS_ihc_group(
        marker,
        marker,
        "Group_median",
        "median",
        ihc_data
      )

    }
  )
)


univ_RFS_continuous_ihc <- bind_rows(
  lapply(
    compartments,
    function(marker) {

      run_cox_RFS_ihc_continuous(
        marker,
        marker,
        ihc_data
      )

    }
  )
)


# ----------------------------------------------------------------------------
# Combine RFS univariate results
# BH correction separately within each threshold
# ----------------------------------------------------------------------------

univ_RFS_ihc_all <- bind_rows(
  univ_RFS_maxstat_ihc,
  univ_RFS_median_ihc,
  univ_RFS_continuous_ihc
) %>%
  group_by(threshold_type) %>%
  mutate(
    p_adj_BH = p.adjust(
      pval,
      method = "BH"
    )
  ) %>%
  ungroup()


# ----------------------------------------------------------------------------
# Format table
# ----------------------------------------------------------------------------

univ_RFS_ihc_formatted <- univ_RFS_ihc_all %>%
  mutate(
    HR_CI = sprintf(
      "%.3f (%.3f–%.3f)",
      HR,
      lower_095,
      upper_095
    ),
    pval = round(
      pval,
      4
    ),
    p_adj_BH = round(
      p_adj_BH,
      4
    )
  ) %>%
  select(
    marker,
    threshold_type,
    HR_CI,
    pval,
    p_adj_BH,
    n,
    n_events,
    n_iter,
    converged
  ) %>%
  arrange(
    marker,
    threshold_type
  )




cutoffs_RFS_cd163 <- make_cutoff_table_ihc(ihc_data, compartments,
                                           "time_recurrence_surv_60m", "recurrence_60m")
univ_RFS_ihc_formatted <- add_cutoffs_ihc(univ_RFS_ihc_formatted, cutoffs_RFS_cd163)
kable(univ_RFS_ihc_formatted, caption = "Univariate Firth Cox regression for RFS: CD163 IHC")

```



```{r}
# ============================================================================
# RFS — SELECT MARKERS FOR MULTIVARIATE
# UNION OF MARKERS WITH p(BH) < 0.10
# ============================================================================

selected_RFS_ihc <- univ_RFS_ihc_all %>%
  filter(
    p_adj_BH < 0.10
  ) %>%
  pull(marker) %>%
  unique()


cat(
  "CD163 markers entering ALL RFS multivariable models:\n"
)

print(selected_RFS_ihc)
```

```{r}
# ============================================================================
# RFS — MULTIVARIATE MAXSTAT / MEDIAN
# adjusted for FIGO + p53
# ============================================================================

perform_cox_RFS_ihc_multiv <- function(
    marker_label,
    marker_var,
    exposure_var,
    data,
    covariates,
    maxit_val = 100000
) {

  df0 <- add_threshold_groups_ihc(
    data,
    "time_recurrence_surv_60m",
    "recurrence_60m",
    marker_var
  )

  df0[[exposure_var]] <- relevel(
    factor(df0[[exposure_var]]),
    ref = "Low"
  )

  df0 <- df0 %>%
    filter(
      !is.na(.data[[exposure_var]]),
      !is.na(recurrence_60m),
      !is.na(time_recurrence_surv_60m),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_recurrence_surv_60m, recurrence_60m) ~ ",
      paste(
        c(
          exposure_var,
          covariates
        ),
        collapse = " + "
      )
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {

    return(
      data.frame(
        marker = marker_label,
        variable = NA_character_,
        coef = NA_real_,
        HR = NA_real_,
        lower_095 = NA_real_,
        upper_095 = NA_real_,
        pval = NA_real_,
        n = nrow(df0),
        n_events = sum(
          df0$recurrence_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE,
        stringsAsFactors = FALSE
      )
    )
  }

  s <- summary(model)

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    variable = names(model$coefficients),
    coef = as.numeric(
      model$coefficients
    ),
    HR = exp(
      as.numeric(
        model$coefficients
      )
    ),
    lower_095 = as.numeric(
      model$ci.lower
    ),
    upper_095 = as.numeric(
      model$ci.upper
    ),
    pval = as.numeric(
      model$prob
    ),
    n = nrow(df0),
    n_events = sum(
      df0$recurrence_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged,
    stringsAsFactors = FALSE
  )
}


# ----------------------------------------------------------------------------
# MAXSTAT
# ----------------------------------------------------------------------------

multiv_RFS_maxstat_ihc <- if (
  length(selected_RFS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_RFS_ihc,
      function(marker) {

        perform_cox_RFS_ihc_multiv(
          marker,
          marker,
          "Group_maxstat",
          ihc_data,
          covariates_full
        )

      }
    )
  ) %>%
    mutate(
      threshold_type = "maxstat",
      .after = marker
    )

} else {

  data.frame()

}


# ----------------------------------------------------------------------------
# MEDIAN
# ----------------------------------------------------------------------------

multiv_RFS_median_ihc <- if (
  length(selected_RFS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_RFS_ihc,
      function(marker) {

        perform_cox_RFS_ihc_multiv(
          marker,
          marker,
          "Group_median",
          ihc_data,
          covariates_full
        )

      }
    )
  ) %>%
    mutate(
      threshold_type = "median",
      .after = marker
    )

} else {

  data.frame()

}
```


```{r}
# ============================================================================
# RFS — MULTIVARIATE CONTINUOUS
# ============================================================================

perform_cox_RFS_ihc_multiv_continuous <- function(
    marker_label,
    marker_var,
    data,
    covariates,
    maxit_val = 100000
) {

  df0 <- data %>%
    filter(
      !is.na(.data[[marker_var]]),
      !is.na(recurrence_60m),
      !is.na(time_recurrence_surv_60m),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_recurrence_surv_60m, recurrence_60m) ~ ",
      marker_var,
      " + ",
      paste(
        covariates,
        collapse = " + "
      )
    )
  )

  model <- tryCatch(
    coxphf(
      fmla,
      data = df0,
      maxstep = 0.0001,
      maxit = maxit_val
    ),
    error = function(e) NULL
  )

  if (is.null(model)) {

    return(
      data.frame(
        marker = marker_label,
        variable = NA_character_,
        coef = NA_real_,
        HR = NA_real_,
        lower_095 = NA_real_,
        upper_095 = NA_real_,
        pval = NA_real_,
        n = nrow(df0),
        n_events = sum(
          df0$recurrence_60m == 1,
          na.rm = TRUE
        ),
        n_iter = NA,
        converged = FALSE,
        stringsAsFactors = FALSE
      )
    )
  }

  s <- summary(model)

  converged <- check_convergence(
    model,
    maxit_val
  )

  data.frame(
    marker = marker_label,
    variable = names(model$coefficients),
    coef = as.numeric(
      model$coefficients
    ),
    HR = exp(
      as.numeric(
        model$coefficients
      )
    ),
    lower_095 = as.numeric(
      model$ci.lower
    ),
    upper_095 = as.numeric(
      model$ci.upper
    ),
    pval = as.numeric(
      model$prob
    ),
    n = nrow(df0),
    n_events = sum(
      df0$recurrence_60m == 1,
      na.rm = TRUE
    ),
    n_iter = paste(
      model$iter,
      collapse = ","
    ),
    converged = converged,
    stringsAsFactors = FALSE
  )
}


multiv_RFS_continuous_ihc <- if (
  length(selected_RFS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_RFS_ihc,
      function(marker) {

        perform_cox_RFS_ihc_multiv_continuous(
          marker,
          marker,
          ihc_data,
          covariates_full
        )

      }
    )
  ) %>%
    mutate(
      threshold_type = "continuous",
      .after = marker
    )

} else {

  data.frame()

}


# ----------------------------------------------------------------------------
# BH correction for continuous marker terms
# ----------------------------------------------------------------------------

if (
  nrow(multiv_RFS_continuous_ihc) > 0
) {

  multiv_RFS_continuous_ihc <-
    multiv_RFS_continuous_ihc %>%
    mutate(
      p_adj_BH = NA_real_
    )

  marker_rows <-
    multiv_RFS_continuous_ihc$variable ==
    multiv_RFS_continuous_ihc$marker

  multiv_RFS_continuous_ihc$p_adj_BH[
    marker_rows
  ] <-
    p.adjust(
      multiv_RFS_continuous_ihc$pval[
        marker_rows
      ],
      method = "BH"
    )
}
```

```{r}
# ============================================================================
# COMBINE RFS MULTIVARIATE RESULTS
# ============================================================================

multiv_RFS_ihc_all <- bind_rows(
  multiv_RFS_maxstat_ihc,
  multiv_RFS_median_ihc,
  multiv_RFS_continuous_ihc
)


# ----------------------------------------------------------------------------
# BH correction separately within threshold type
# ----------------------------------------------------------------------------

multiv_RFS_ihc_all <-
  multiv_RFS_ihc_all %>%
  group_by(
    threshold_type
  ) %>%
  mutate(
    marker_test =
      grepl(
        "^Group_",
        variable
      ) |
      variable == marker,

    p_adj_BH =
      ifelse(
        marker_test,

        p.adjust(
          pval[
            marker_test
          ],
          method = "BH"
        )[match(
          pval,
          pval[
            marker_test
          ]
        )],

        NA_real_
      )
  ) %>%
  ungroup()


# ----------------------------------------------------------------------------
# Format table
# ----------------------------------------------------------------------------

format_table_RFS_ihc <- function(df) {

  df %>%
    mutate(

      variable_label = case_when(

        grepl(
          "^Group_maxstat",
          variable
        ) ~
          "Enrichment (High vs Low)",

        grepl(
          "^Group_median",
          variable
        ) ~
          "Enrichment (High vs Low)",

        variable == marker ~
          "Continuous (per unit)",

        variable ==
          "figo_ini_ava0 (Initial)" ~
          "FIGO (Initial)",

        variable ==
          "p53_ihc_statusWt" ~
          "p53 (Normal)",

        TRUE ~
          variable
      ),

      HR_CI = sprintf(
        "%.4f (%.4f–%.4f)",
        HR,
        lower_095,
        upper_095
      ),

      coef = round(
        coef,
        4
      ),

      pval = round(
        pval,
        4
      ),

      p_adj_BH = round(
        p_adj_BH,
        4
      )
    ) %>%
    select(
      marker,
      threshold_type,
      Variable = variable_label,
      Coef = coef,
      `HR (95% CI)` = HR_CI,
      `p-value` = pval,
      `p (BH-adj)` = p_adj_BH,
      n,
      n_events,
      n_iter,
      converged
    )
}


multiv_RFS_ihc_formatted <-
  format_table_RFS_ihc(
    multiv_RFS_ihc_all
  ) %>%
  arrange(
    marker,
    threshold_type
  )


multiv_RFS_ihc_formatted <- add_cutoffs_ihc(multiv_RFS_ihc_formatted, cutoffs_RFS_cd163)


kable(
  multiv_RFS_ihc_formatted,
  caption =
    "Multivariate Firth Cox regression for RFS: CD163 IHC"
)
```


```{r}
# ============================================================================
# RFS — PROPORTIONAL HAZARDS ASSUMPTION
# Schoenfeld residuals
# ============================================================================

check_ph_RFS_ihc <- function(
    marker_label,
    marker_var,
    data,
    exposure_var,
    covariates
) {

  df0 <- add_threshold_groups_ihc(
    data,
    "time_recurrence_surv_60m",
    "recurrence_60m",
    marker_var
  )

  df0[[exposure_var]] <- relevel(
    factor(
      df0[[exposure_var]]
    ),
    ref = "Low"
  )

  d <- df0 %>%
    mutate(
      recurrence_60m =
        as.numeric(
          as.character(
            recurrence_60m
          )
        )
    ) %>%
    filter(
      !is.na(.data[[exposure_var]]),
      !is.na(
        time_recurrence_surv_60m
      ),
      !is.na(
        recurrence_60m
      ),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_recurrence_surv_60m, recurrence_60m) ~ ",
      paste(
        c(
          exposure_var,
          covariates
        ),
        collapse = " + "
      )
    )
  )

  fit <- tryCatch(
    coxph(
      fmla,
      data = d
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) {

    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$recurrence_60m == 1,
          na.rm = TRUE
        ),
        note = "coxph_fit_failed"
      )
    )
  }

  zph <- tryCatch(
    cox.zph(fit),
    error = function(e) NULL
  )

  if (is.null(zph)) {

    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$recurrence_60m == 1,
          na.rm = TRUE
        ),
        note = "zph_failed"
      )
    )
  }

  tab <- as.data.frame(
    zph$table
  )

  tab$term <- rownames(
    zph$table
  )

  global_p <-
    tab$p[
      tab$term == "GLOBAL"
    ]

  data.frame(
    marker = marker_label,
    term = tab$term,
    chisq = round(
      tab$chisq,
      3
    ),
    df = tab$df,
    p = round(
      tab$p,
      4
    ),
    global_p = round(
      global_p,
      4
    ),
    n = nrow(d),
    n_events = sum(
      d$recurrence_60m == 1,
      na.rm = TRUE
    ),
    note = NA_character_
  )
}


# ----------------------------------------------------------------------------
# PH — MAXSTAT
# ----------------------------------------------------------------------------

ph_RFS_maxstat_ihc <- if (
  length(selected_RFS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_RFS_ihc,
      function(m) {

        check_ph_RFS_ihc(
          m,
          m,
          ihc_data,
          "Group_maxstat",
          covariates_full
        )

      }
    )
  ) %>%
    mutate(
      threshold_type = "maxstat",
      .after = marker
    )

} else {

  data.frame()

}


# ----------------------------------------------------------------------------
# PH — MEDIAN
# ----------------------------------------------------------------------------

ph_RFS_median_ihc <- if (
  length(selected_RFS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_RFS_ihc,
      function(m) {

        check_ph_RFS_ihc(
          m,
          m,
          ihc_data,
          "Group_median",
          covariates_full
        )

      }
    )
  ) %>%
    mutate(
      threshold_type = "median",
      .after = marker
    )

} else {

  data.frame()

}
```



```{r}
# ============================================================================
# RFS — PH TEST FOR CONTINUOUS MODELS
# ============================================================================

check_ph_RFS_ihc_continuous <- function(
    marker_label,
    marker_var,
    data,
    covariates
) {

  d <- data %>%
    filter(
      !is.na(
        .data[[marker_var]]
      ),
      !is.na(
        time_recurrence_surv_60m
      ),
      !is.na(
        recurrence_60m
      ),
      if_all(
        all_of(covariates),
        ~ !is.na(.)
      )
    ) %>%
    mutate(
      recurrence_60m =
        as.numeric(
          as.character(
            recurrence_60m
          )
        )
    ) %>%
    droplevels()

  fmla <- as.formula(
    paste0(
      "Surv(time_recurrence_surv_60m, recurrence_60m) ~ ",
      marker_var,
      " + ",
      paste(
        covariates,
        collapse = " + "
      )
    )
  )

  fit <- tryCatch(
    coxph(
      fmla,
      data = d
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) {

    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$recurrence_60m == 1,
          na.rm = TRUE
        ),
        note = "coxph_fit_failed"
      )
    )
  }

  zph <- tryCatch(
    cox.zph(fit),
    error = function(e) NULL
  )

  if (is.null(zph)) {

    return(
      data.frame(
        marker = marker_label,
        term = NA,
        chisq = NA,
        df = NA,
        p = NA,
        global_p = NA,
        n = nrow(d),
        n_events = sum(
          d$recurrence_60m == 1,
          na.rm = TRUE
        ),
        note = "zph_failed"
      )
    )
  }

  tab <- as.data.frame(
    zph$table
  )

  tab$term <- rownames(
    zph$table
  )

  global_p <-
    tab$p[
      tab$term == "GLOBAL"
    ]

  data.frame(
    marker = marker_label,
    term = tab$term,
    chisq = round(
      tab$chisq,
      3
    ),
    df = tab$df,
    p = round(
      tab$p,
      4
    ),
    global_p = round(
      global_p,
      4
    ),
    n = nrow(d),
    n_events = sum(
      d$recurrence_60m == 1,
      na.rm = TRUE
    ),
    note = NA_character_
  )
}


ph_RFS_continuous_ihc <- if (
  length(selected_RFS_ihc) > 0
) {

  bind_rows(
    lapply(
      selected_RFS_ihc,
      function(m) {

        check_ph_RFS_ihc_continuous(
          m,
          m,
          ihc_data,
          covariates_full
        )

      }
    )
  ) %>%
    mutate(
      threshold_type = "continuous",
      .after = marker
    )

} else {

  data.frame()

}


# ----------------------------------------------------------------------------
# Combine PH results
# ----------------------------------------------------------------------------

ph_RFS_ihc_all <- bind_rows(
  ph_RFS_maxstat_ihc,
  ph_RFS_median_ihc,
  ph_RFS_continuous_ihc
) %>%
  arrange(
    marker,
    threshold_type,
    term
  )


kable(
  ph_RFS_ihc_all,
  caption =
    "Proportional hazards assumption for multivariable RFS models: CD163 IHC"
)
```


```{r}
# ============================================================================
# RFS — SELECT MULTIVARIABLE MODELS FOR BOOTSTRAP
# BH-adjusted p < 0.05
# ============================================================================

sig_maxstat_RFS_ihc <- multiv_RFS_ihc_all %>%
  filter(
    threshold_type == "maxstat",
    grepl(
      "^Group_maxstat",
      variable
    ),
    !is.na(p_adj_BH),
    p_adj_BH < 0.05
  ) %>%
  pull(marker) %>%
  unique()


sig_median_RFS_ihc <- multiv_RFS_ihc_all %>%
  filter(
    threshold_type == "median",
    grepl(
      "^Group_median",
      variable
    ),
    !is.na(p_adj_BH),
    p_adj_BH < 0.05
  ) %>%
  pull(marker) %>%
  unique()


sig_continuous_RFS_ihc <- multiv_RFS_ihc_all %>%
  filter(
    threshold_type == "continuous",
    variable == marker,
    !is.na(p_adj_BH),
    p_adj_BH < 0.05
  ) %>%
  pull(marker) %>%
  unique()


cat(
  "RFS maxstat models for bootstrap:\n"
)

print(
  sig_maxstat_RFS_ihc
)


cat(
  "\nRFS median models for bootstrap:\n"
)

print(
  sig_median_RFS_ihc
)


cat(
  "\nRFS continuous models for bootstrap:\n"
)

print(
  sig_continuous_RFS_ihc
)
```
#There are no padj<0.05 in multivariate models so i do not need to run bootsrap for these






```{r}
relabel <- function(df) df %>% mutate(marker = recode(marker,
  "Positive_cells_tumoral_area"       = "Positive_cells_overall_tumoral_area",
  "Positive_cells_tumor_stroma"       = "Positive_cells_tumor_associated_stroma",
  "Positive_cells_perivascular_areas" = "Positive_cells_perivascular_areas"))

tabs <- list(
  Uni_recurrence_CD163   = relabel(univ_RFS_ihc_formatted),
  Multi_recurrence_CD163 = relabel(multiv_RFS_ihc_formatted),
  Uni_death_CD163        = relabel(univ_DSS_ihc_formatted),
  Multi_death_CD163      = relabel(multiv_DSS_ihc_formatted)
)
wb <- createWorkbook()
for (nm in names(tabs)) {
  addWorksheet(wb, nm)
  writeData(wb, nm, tabs[[nm]], headerStyle = createStyle(textDecoration = "bold"))
  setColWidths(wb, nm, cols = 1:ncol(tabs[[nm]]), widths = "auto")
}
saveWorkbook(wb, "CD163_survival_with_cutoffs.xlsx", overwrite = TRUE)
```




```{r}
   wb9 <- createWorkbook()
   addWorksheet(wb9, "Bootstrap_CD163_death")
   writeData(wb9, "Bootstrap_CD163_death", relabel(boot_DSS_ihc_all),
             headerStyle = createStyle(textDecoration = "bold"))
   addWorksheet(wb9, "Bootstrap_CD163_recurrence")
   writeData(wb9, "Bootstrap_CD163_recurrence",
             data.frame(Note = "No CD163 RFS multivariable model reached BH-adjusted p < 0.05; bootstrap not performed."))
   saveWorkbook(wb9, "CD163_bootstrap.xlsx", overwrite = TRUE)
```







































#PD-L1 CD163 interaction analysis

```{r}
# ============================================================
# 1. DEFINE CD163 HIGH/LOW USING THE MAXSTAT THRESHOLD
#    (overall tumor area, DSS)
# ============================================================

# Find the optimal CD163 tumor-area cutoff using maxstat
cd163_maxstat_cutoff <- find_maxstat_cutoff_ihc(
  data = ihc_data,
  marker_var = "Positive_cells_tumoral_area",
  time_var = "time_death_surv_60m",
  event_var = "death_disease_60m"
)

cd163_maxstat_cutoff


# Create CD163 High/Low using the SAME maxstat cutoff
# for ALL subsequent analyses

ihc_data <- ihc_data %>%
  mutate(
    CD163_group = case_when(
      is.na(Positive_cells_tumoral_area) ~ NA_character_,
      Positive_cells_tumoral_area > cd163_maxstat_cutoff ~ "High",
      Positive_cells_tumoral_area <= cd163_maxstat_cutoff ~ "Low"
    ),

    CD163_group = factor(
      CD163_group,
      levels = c("Low", "High")
    ),

    PDL1_group = factor(
      pdl1_immune_cells_pos_category,
      levels = c("Negative", "Positive")
    )
  )


# Check distribution

table(
  ihc_data$CD163_group,
  useNA = "ifany"
)

table(
  ihc_data$PDL1_group,
  useNA = "ifany"
)


# ============================================================
# 2. CREATE THE FOUR COMBINED GROUPS
# ============================================================

ihc_data <- ihc_data %>%
  mutate(
    immuno_group = case_when(
      PDL1_group == "Negative" & CD163_group == "Low" ~
        "PD-L1- CD163 low",

      PDL1_group == "Negative" & CD163_group == "High" ~
        "PD-L1- CD163 high",

      PDL1_group == "Positive" & CD163_group == "Low" ~
        "PD-L1+ CD163 low",

      PDL1_group == "Positive" & CD163_group == "High" ~
        "PD-L1+ CD163 high",

      TRUE ~ NA_character_
    ),

    immuno_group = factor(
      immuno_group,
      levels = c(
        "PD-L1- CD163 low",
        "PD-L1- CD163 high",
        "PD-L1+ CD163 low",
        "PD-L1+ CD163 high"
      )
    )
  )


# ============================================================
# 3. CHECK NUMBERS IN THE FOUR GROUPS
# ============================================================

table(
  ihc_data$immuno_group,
  useNA = "ifany"
)


# Show the actual maxstat cutoff and group sizes
cat(
  "\nCD163 tumoral area maxstat cutoff:",
  cd163_maxstat_cutoff,
  "\n"
)

print(
  table(
    ihc_data$PDL1_group,
    ihc_data$CD163_group,
    useNA = "ifany"
  )
)
```



```{r}
# ============================================================
# ASSOCIATION BETWEEN PD-L1 STATUS AND CD163 GROUP
# Fisher's exact test + concordance
# ============================================================

tab_pdl1_cd163 <- table(
  PDL1  = ihc_data$PDL1_group,
  CD163 = ihc_data$CD163_group
)

tab_pdl1_cd163

fisher_pdl1_cd163 <- fisher.test(tab_pdl1_cd163)
fisher_pdl1_cd163

# CD163-high proportion within each PD-L1 group
prop.table(tab_pdl1_cd163, margin = 1)

# Concordance: PD-L1-/CD163 low + PD-L1+/CD163 high
n_concordant <- tab_pdl1_cd163["Negative", "Low"] +
  tab_pdl1_cd163["Positive", "High"]

n_total <- sum(tab_pdl1_cd163)

cat(
  "Concordant:", n_concordant, "/", n_total,
  sprintf("(%.1f%%)", 100 * n_concordant / n_total), "\n",
  "Discordant:", n_total - n_concordant, "/", n_total,
  sprintf("(%.1f%%)", 100 * (n_total - n_concordant) / n_total), "\n"
)
```




```{r}
# ============================================================
# 4-GROUP KAPLAN-MEIER: PD-L1 × CD163
# ============================================================

ihc_data$immuno_group <- factor(
  ihc_data$immuno_group,
  levels = c(
    "PD-L1- CD163 low",
    "PD-L1- CD163 high",
    "PD-L1+ CD163 low",
    "PD-L1+ CD163 high"
  )
)

# ------------------------------------------------------------
# Survival object
# ------------------------------------------------------------

surv_obj <- Surv(
  time = ihc_data$time_death_surv_60m,
  event = as.numeric(
    as.character(ihc_data$death_disease_60m)
  )
)

# ------------------------------------------------------------
# Fit KM model
# ------------------------------------------------------------

km_fit <- survfit(
  surv_obj ~ immuno_group,
  data = ihc_data
)

# ------------------------------------------------------------
# Colors: light blue = PD-L1-, dark blue = PD-L1+
# ------------------------------------------------------------

custom_colors <- c(
   "#6baed6",  # PD-L1- CD163 low
   "#6baed5",  # PD-L1- CD163 high
  "#08519d",  # PD-L1+ CD163 low
  "#08519c"   # PD-L1+ CD163 high
)

# Dashed = CD163 low
# Solid  = CD163 high

custom_linetypes <- c(
  "dashed",
  "solid",
  "dashed",
  "solid"
)

# ------------------------------------------------------------
# Plot
# ------------------------------------------------------------

p <- ggsurvplot(
  km_fit,
  data = ihc_data,
  pval = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.3,
  xlab = "Time to Event (Months)",
  ylab = "Survival Probability",
  break.time.by = 10,
  palette = custom_colors,
  linetype = custom_linetypes,
  legend.title = "Immuno Group",
  legend.labs = levels(ihc_data$immuno_group),
  legend = c(0.8, 0.2),
  title = "Kaplan-Meier Survival Curve by PD-L1 and CD163 Status"
)

print(p)
```



```{r}
# ============================================================
# 5. FOUR-GROUP SUMMARY: N AND EVENTS
# ============================================================

four_group_summary <- data.frame(
  Group = levels(ihc_data$immuno_group)
)

four_group_summary$N <- as.numeric(
  table(ihc_data$immuno_group)
)

four_group_summary$Events <- as.numeric(
  tapply(
    as.numeric(as.character(ihc_data$death_disease_60m)),
    ihc_data$immuno_group,
    sum,
    na.rm = TRUE
  )
)

four_group_summary
```



```{r}
# ============================================================
# 6. OVERALL FOUR-GROUP LOG-RANK TEST
# ============================================================

logrank_4group <- survdiff(
  surv_obj ~ immuno_group,
  data = ihc_data
)

logrank_4group_p <- pchisq(
  logrank_4group$chisq,
  df = length(logrank_4group$n) - 1,
  lower.tail = FALSE
)

cat(
  "\nOverall 4-group log-rank test:\n",
  "Chi-square =", round(logrank_4group$chisq, 3),
  "\nDegrees of freedom =", length(logrank_4group$n) - 1,
  "\np =", signif(logrank_4group_p, 3),
  "\n"
)
```



```{r}
# ============================================================
# 7. EXPLORATORY FIRTH COX MODEL: HRs, 95% CIs, N AND EVENTS
#    Reference group: PD-L1- CD163 low
#    (coxphf CIs are already on the HR scale -> no exp())
# ============================================================

ihc_data$immuno_group <- relevel(
  ihc_data$immuno_group,
  ref = "PD-L1- CD163 low"
)

cox_4group_firth <- coxphf(
  surv_obj ~ immuno_group,
  data = ihc_data,
  maxstep = 0.0001,
  maxit = 100000
)

converged_4group <- check_convergence(cox_4group_firth, 100000)

group_N <- table(ihc_data$immuno_group)

group_events <- tapply(
  as.numeric(as.character(ihc_data$death_disease_60m)),
  ihc_data$immuno_group,
  sum,
  na.rm = TRUE
)

cox_4group_results <- data.frame(
  Group = levels(ihc_data$immuno_group),
  N = as.numeric(group_N),
  Events = as.numeric(group_events),
  coef = c(0, cox_4group_firth$coefficients),
  HR = c(1, exp(cox_4group_firth$coefficients)),
  Lower_95CI = c(NA, cox_4group_firth$ci.lower),
  Upper_95CI = c(NA, cox_4group_firth$ci.upper),
  P_value = c(NA, cox_4group_firth$prob),
  converged = converged_4group,
  row.names = NULL
)

cox_4group_results
```



##Interaction pdl1 and CD163+cells
```{r}
# ============================================================
# 8. FORMAL CD163 × PD-L1 INTERACTION ANALYSIS
#    Firth-penalized Cox proportional hazards model
# ============================================================

# ------------------------------------------------------------
# Prepare variables
# ------------------------------------------------------------

interaction_data <- ihc_data[
  ihc_data$immuno_group %in% c(
    "PD-L1- CD163 low",
    "PD-L1- CD163 high",
    "PD-L1+ CD163 low",
    "PD-L1+ CD163 high"
  ),
]

# Create PD-L1 variable
interaction_data$PDL1 <- ifelse(
  grepl("PD-L1\\+", interaction_data$immuno_group),
  "Positive",
  "Negative"
)

interaction_data$PDL1 <- factor(
  interaction_data$PDL1,
  levels = c("Negative", "Positive")
)

# Create CD163 variable
interaction_data$CD163 <- ifelse(
  grepl("CD163 high", interaction_data$immuno_group),
  "High",
  "Low"
)

interaction_data$CD163 <- factor(
  interaction_data$CD163,
  levels = c("Low", "High")
)

# ------------------------------------------------------------
# Create binary event variable
# death_disease_60m is coded as 0/1
# ------------------------------------------------------------

interaction_data$event <- as.numeric(
  as.character(interaction_data$death_disease_60m)
)

# ------------------------------------------------------------
# Remove incomplete observations
# ------------------------------------------------------------

interaction_data <- interaction_data[
  !is.na(interaction_data$time_death_surv_60m) &
  !is.na(interaction_data$event) &
  !is.na(interaction_data$PDL1) &
  !is.na(interaction_data$CD163),
]

# Check numbers
table(interaction_data$PDL1, interaction_data$CD163)
```

```{r}
# ============================================================
# Firth Cox model WITHOUT interaction (additive model)
# ============================================================

cox_no_interaction <- coxphf(
  Surv(time_death_surv_60m, event) ~
    CD163 + PDL1,
  data = interaction_data,
  maxstep = 0.0001,
  maxit = 100000
)

converged_no_int <- check_convergence(cox_no_interaction, 100000)

cox_no_interaction
converged_no_int

cox_additive_results <- data.frame(
  Variable = names(cox_no_interaction$coefficients),
  HR = exp(cox_no_interaction$coefficients),
  Lower_95CI = cox_no_interaction$ci.lower,
  Upper_95CI = cox_no_interaction$ci.upper,
  P_value = cox_no_interaction$prob,
  converged = converged_no_int,
  row.names = NULL
)

cox_additive_results
```

```{r}
# ============================================================
# Firth Cox model WITH CD163 × PD-L1 interaction
# ============================================================

cox_interaction <- coxphf(
  Surv(time_death_surv_60m, event) ~
    CD163 * PDL1,
  data = interaction_data,
  maxstep = 0.0001,
  maxit = 100000
)

converged_int <- check_convergence(cox_interaction, 100000)

cox_interaction
converged_int

cox_results <- data.frame(
  Variable = names(cox_interaction$coefficients),
  HR = exp(cox_interaction$coefficients),
  Lower_95CI = cox_interaction$ci.lower,
  Upper_95CI = cox_interaction$ci.upper,
  P_value = cox_interaction$prob,
  converged = converged_int,
  row.names = NULL
)

cox_results
```



#Does adding the CD163 × PD-L1 interaction significantly improve the Cox model?
```{r}
# ============================================================
# Penalized likelihood ratio test for the CD163 x PD-L1 interaction
# coxphftest() compares the full model with the model without the
# interaction term, using the Firth penalty of the full model
# (the manual comparison of penalized log-likelihoods is not valid).
# ============================================================

lrt_interaction <- coxphftest(
  Surv(time_death_surv_60m, event) ~ CD163 * PDL1,
  test = ~ CD163:PDL1,
  data = interaction_data
)

lrt_interaction
```



```{r}
# ============================================================
# Proportional hazards assumption
#
# cox.zph() requires a plain coxph object -- coxphf (Firth) has no
# residuals/zph method. This fits parallel unpenalized coxph()
# models with the identical formulas purely for this diagnostic.
# Effect sizes / inference come from the Firth models above.
# ============================================================

# Additive model
cox_additive_ph <- coxph(
  Surv(time_death_surv_60m, event) ~
    CD163 + PDL1,
  data = interaction_data
)

ph_test_additive <- cox.zph(cox_additive_ph)
ph_test_additive

# Interaction model
cox_interaction_ph <- coxph(
  Surv(time_death_surv_60m, event) ~
    CD163 * PDL1,
  data = interaction_data
)

ph_test <- cox.zph(cox_interaction_ph)

ph_test

plot(ph_test)
```



```{r}
# ============================================================
# EXPORT - Supplementary Table 9
# ============================================================

ph_to_df <- function(zph, model_label) {
  data.frame(
    model = model_label,
    term = rownames(zph$table),
    chisq = round(zph$table[, "chisq"], 3),
    df = zph$table[, "df"],
    p = round(zph$table[, "p"], 4),
    row.names = NULL
  )
}

st9 <- list(
  PDL1_CD163_association = data.frame(
    PDL1 = rownames(tab_pdl1_cd163),
    CD163_low = as.numeric(tab_pdl1_cd163[, "Low"]),
    CD163_high = as.numeric(tab_pdl1_cd163[, "High"]),
    Fisher_p = c(fisher_pdl1_cd163$p.value, NA),
    Cutoff_tumor_area = c(paste0(">", cd163_maxstat_cutoff, "%"), NA)
  ),
  Four_groups = cox_4group_results,
  Additive_model = cox_additive_results,
  Interaction_model = cox_results,
  PH_tests = rbind(
    ph_to_df(ph_test_additive, "additive"),
    ph_to_df(ph_test, "interaction")
  )
)

wb <- createWorkbook()
for (nm in names(st9)) {
  addWorksheet(wb, nm)
  writeData(wb, nm, st9[[nm]], headerStyle = createStyle(textDecoration = "bold"))
  setColWidths(wb, nm, cols = 1:ncol(st9[[nm]]), widths = "auto")
}
saveWorkbook(wb, "Supplementary_Table_9_PDL1_CD163.xlsx", overwrite = TRUE)
```


