

# **NORMALITY TESTS**
#We perform Anderson-Darling tests to examine the distribution of the numerical variables

# Install the packages if not already installed
```{r packages}
library(gtsummary)
library(magrittr)
library(nortest)
library(knitr)
library(dplyr)
```


# Normality in continuous variables, we choose Ad test because we have a datset n>50: "For larger datasets (n>50), other tests like the Kolmogorov-Smirnov test, Anderson-Darling test, or Lilliefors test may be more robust alternatives".
```{r}
numeric_data <- select_if(TP53status_clinical, is.numeric)
print(colnames(numeric_data))

#We define the numeric columns to test for normality
columns_to_test <- colnames(numeric_data)

#We create an empty dataframe to store results
results_df <- data.frame(Column = character(),
                         Test_utilizado = character(),
                         Valor_p = numeric(),
                         stringsAsFactors = FALSE)

#We define a loop to check normality in the previously defined columns
for (col in columns_to_test) {
  #Test Anderson-Darling:
  ad_test <- ad.test(TP53status_clinical[[col]])
  
  #Interpretation
  if (ad_test$p.value < 0.05) {
    normality <- "Distribución no normal"
  } else {
    normality <- "Distribución normal"
  }
  
  #We add the results into the created empty dataframe
  results_df <- rbind(results_df, data.frame(Variable = col,
                                             Test_utilizado= "Anderson-Darling",
                                             Valor_p = round(ad_test$p.value, digits = 3),
                                             Interpretación = normality))
}

#We print the results
knitr::kable(results_df, caption = "Resumen estadísticos normalidad")
```


#As we see that all variables are non-normal, we will perform Mann–Whitney U test in all the comparisons for numeric variables. Mann–Whitney U test, also called Mann–Whitney–Wilcoxon (MWW/MWU), Wilcoxon rank-sum test, or Wilcoxon–Mann–Whitney test) is a nonparametric test appropiate for non-normal variables.



# **DESCRIPTIVE TABLES**
```{r}
#Select the variables of interest
descriptive_HTG <- TP53status_clinical %>% select(
    "hpv_status",
    "age",
    "figo_ini_ava",
    "tumor_size",
    "tumor_size_category",
    "depth_invasion",
    "depth_invasion_category",
    "hist_type_category",
    "vli", #Vascular lymphatic invasion (VLI)
    "nodal_status",#Lymph node status
    "margins",
    "radiotherapy",
    "surgery_type_category",
    "p53_ihc_status",
    "time_death_surv_60m", #this is time until lastfollow up up with a threshold of 60m
    "death_disease_60m",
    "time_recurrence_surv_60m",
    "recurrence_60m",
    "time_last_followup"
    )
```






#We create a summary table, we do fisher test for categorical as we know that most of the variables contain less than 5 observations in one of the groups
```{r}
library(gtsummary)
library(webshot)
library(dplyr)
library(remotes)

descriptive_table_HTG <- descriptive_HTG %>%
  tbl_summary(
    by = "hpv_status",  # Quote the column name explicitly
    statistic = list(
      all_categorical() ~ "{n} / {N} ({p}%)",  # Ensure counts and percentages are formatted properly
      all_continuous() ~ "{median} ({p25}, {p75})"  # Display medians and percentiles for continuous variables
    ),
    type = list(  # Explicitly set categorical variables
      pdl1_immune_cells_pos ~ "continuous",  # Explicitly define as continuous
      figo_ini_ava ~ "categorical",
      tumor_size_category ~ "categorical",
      depth_invasion_category ~ "categorical",
      hist_type_category ~ "categorical",
      vli ~ "categorical",
      nodal_status ~ "categorical",
      margins ~ "categorical",
      radiotherapy ~ "categorical",
      surgery_type_category ~ "categorical",
      p53_ihc_status ~ "categorical",
      recurrence_60m ~ "categorical",  # Ensure recurrence_60m is treated as categorical
      death_disease_60m ~ "categorical"  # Ensure death_disease_60m is treated as categorical
    ),
    digits = list(
      all_categorical() ~ c(0, 0, 1),  # Set counts (n, N) as integers and percentages (p) with one decimal
      all_continuous() ~ 1  # One decimal for continuous variables
    )
  ) %>%
add_p(
    test = list(
      all_categorical() ~ function(data, variable, by, ...) {
        # Fisher test with simulated p-value
        result <- fisher.test(data[[variable]], data[[by]], simulate.p.value = TRUE)
        tibble::tibble(p.value = result$p.value)  # Return as tibble
      },
      all_continuous() ~ function(data, variable, by, ...) {
        # Wilcoxon test with approximate p-value
        result <- wilcox.test(data[[variable]] ~ data[[by]], exact = FALSE)
        tibble::tibble(p.value = result$p.value)  # Return as tibble
      }
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  )
descriptive_table_HTG
```




#We create a summary table within only HPV-Independent group by p53 ihc status.
```{r}
library(gtsummary)
library(dplyr)

# 1️⃣ Filter dataset to HPV-negative cases only
descriptive_HTG_neg <- descriptive_HTG %>%
  filter(hpv_status == "Negative")

# 2️⃣ Create summary table comparing p53_ihc_status groups
descriptive_table_HTG_neg <- descriptive_HTG_neg %>%
  tbl_summary(
    by = "p53_ihc_status",  # compare Wt vs Mutant
    statistic = list(
      all_categorical() ~ "{n} / {N} ({p}%)",
      all_continuous() ~ "{median} ({p25}, {p75})"
    ),
    type = list(
      pdl1_immune_cells_pos ~ "continuous",
      figo_ini_ava ~ "categorical",
      tumor_size_category ~ "categorical",
      depth_invasion_category ~ "categorical",
      hist_type_category ~ "categorical",
      vli ~ "categorical",
      nodal_status ~ "categorical",
      margins ~ "categorical",
      radiotherapy ~ "categorical",
      surgery_type_category ~ "categorical",
      recurrence_60m ~ "categorical",
      death_disease_60m ~ "categorical"
    ),
    digits = list(
      all_categorical() ~ c(0, 0, 1),
      all_continuous() ~ 1
    )
  ) %>%
  add_p(
    test = list(
      all_categorical() ~ function(data, variable, by, ...) {
        result <- fisher.test(data[[variable]], data[[by]], simulate.p.value = TRUE)
        tibble::tibble(p.value = result$p.value)
      },
      all_continuous() ~ function(data, variable, by, ...) {
        result <- wilcox.test(data[[variable]] ~ data[[by]], exact = FALSE)
        tibble::tibble(p.value = result$p.value)
      }
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  )

# 3️⃣ Show the final summary table
descriptive_table_HTG_neg

```






