
#XCELL in all cases by HPV status

```{r}
library(coxphf)
library(knitr)
library(openxlsx)
library(immunedeconv)
library(tidyr)
library(tibble)
library(survival)
library(dplyr)
library(survminer)
library(ggrepel)
library(grid)
library(scales) 
```


```{r}
xcell <- deconvolute(counts_filtered_filtered, "xcell")
length(xcell$cell_type)

#We only obtain 39 cell types. This function automatically drops any cell type that has zero (or near-zero) signal across all samples, that's why we're seeing only 39 out of 64.

xcell_long <- xcell %>%
  pivot_longer(
    cols = -cell_type, # All columns except `cell_type`
    names_to = "Sample", 
    values_to = "Fraction"
  )

analysis_data <- xcell_long %>%
  left_join(TP53status_clinical, by = c( "Sample"="id"))
```

```{r}
threshold <- 0.01      # fraction cutoff
min_presence <- 0.1   # 10% of samples must exceed threshold

 
# Cell types to manually exclude
scores_exclude <- c("microenvironment score", "immune score", "stroma score")

# Keep original data
analysis_data_unfiltered <- analysis_data %>%
  filter(!(cell_type %in% scores_exclude))


# Calculate proportion present
present_summary <- analysis_data_unfiltered %>%
  group_by(cell_type) %>%
  summarise(present = mean(Fraction > threshold), .groups = "drop")

# Filter cell types that meet criteria
filtered_cell_types <- present_summary %>%
  filter(present >= min_presence) %>%
  pull(cell_type)

# Get excluded cell types BEFORE filtering
excluded_cell_types <- setdiff(unique(analysis_data_unfiltered$cell_type),
                               filtered_cell_types)

print(excluded_cell_types)

# Keep only filtered types
analysis_data <- analysis_data_unfiltered %>%
  filter(cell_type %in% filtered_cell_types)

# Plot excluded ones if needed
excluded_data <- analysis_data_unfiltered %>%
  filter(cell_type %in% excluded_cell_types)

if(nrow(excluded_data) > 0) {
  p <- excluded_data %>%
    ggplot(aes(x = Fraction)) +
    geom_histogram(bins = 30, fill = "tomato", color = "black") +
    facet_wrap(~ cell_type, scales = "free_x") +
    theme_minimal() +
    labs(
      title = "Histogram of Fraction for Excluded Cell Types",
      x = "Fraction",
      y = "Count"
    )
  print(p)
} else {
  message("No cell types excluded — try adjusting threshold or min_presence.")
}




p<-analysis_data %>%
  ggplot(aes(x = Fraction)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~ cell_type, scales = "free_x") +
  theme_minimal() +
  labs(title = "Distribution of Cell Fractions Across Cell Types",
       x = "Fraction",
       y = "Count")
p

length(unique(analysis_data$cell_type))
#We detect 39 and we keep 28
```








#together the included and excluded:
```{r}
# Combine included and excluded into one dataset
plot_data <- analysis_data_unfiltered %>%
  mutate(status = ifelse(cell_type %in% filtered_cell_types, "Included", "Excluded"))

# Plot with different colors
p <- plot_data %>%
  ggplot(aes(x = Fraction, fill = status)) +
  geom_histogram(bins = 30, color = "black") +
  facet_wrap(~ cell_type, scales = "fixed") +
  scale_fill_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Distribution of Cell Fractions Across Cell Types",
    x = "Fraction",
    y = "Count",
    fill = "Status"
  )

print(p)




p <- plot_data %>%
  ggplot(aes(x = Fraction, color = status, fill = status)) +
  geom_density(alpha = 0.3) +  # semi-transparent fill
  facet_wrap(~ cell_type, scales = "fixed") +
  scale_fill_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  scale_color_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Density of Cell Fractions Across Cell Types",
    x = "Fraction",
    y = "Density",
    fill = "Status",
    color = "Status"
  )
print(p)




library(dplyr)
library(ggplot2)
library(forcats)

# Reorder cell_type by mean Fraction
plot_data <- plot_data %>%
  group_by(cell_type) %>%
  mutate(mean_fraction = mean(Fraction)) %>%
  ungroup() %>%
  mutate(cell_type = fct_reorder(cell_type, mean_fraction))

# Plot
p <- plot_data %>%
  ggplot(aes(x = Fraction, y = cell_type, color = status)) +
  geom_point(position = position_jitter(height = 0.2, width = 0), alpha = 0.7) +
  scale_color_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Cell Fractions Across Cell Types (Ordered by Mean Fraction)",
    x = "Fraction",
    y = "Cell Type",
    color = "Status"
  )

print(p)
```



```{r}
# --- Step 1: Filter out progenitor cell types
exclude_celltypes <- c(
"Common lymphoid progenitor",     
"Common myeloid progenitor", 
"Granulocyte-monocyte progenitor")

analysis_data <- analysis_data %>%
  filter(!cell_type %in% exclude_celltypes)

length(unique(analysis_data$cell_type))
#We keep 25
```




```{r}
# Load necessary library
library(RColorBrewer)

# Create an extended color palette
n_categories <- length(unique(xcell$cell_type))  # Calculate the number of unique cell types
extended_palette <- colorRampPalette(brewer.pal(12, "Paired"))(n_categories)  # Extend the Paired palette


p <- analysis_data %>%
  ggplot(aes(x = Sample, y = Fraction, fill = cell_type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = extended_palette) +
  theme_minimal() +
  labs(
    title = "XCELL Cell Fractions (Filtered Cell Types)",
    x = "Sample",
    y = "Fraction",
    fill = "Cell Type"
  )

print(p)
```











#Together the included and excluded:
```{r}
# Combine included and excluded into one dataset, and have all info
# plot_data <- analysis_data_unfiltered %>%
#   mutate(status = case_when(
#     cell_type %in% filtered_cell_types ~ "Included",
#     cell_type %in% exclude_celltypes   ~ "Excluded (Progenitor)",
#     TRUE                               ~ "Excluded / Not detected"
#   ))

#To remove it from the plot
plot_data <- analysis_data_unfiltered %>%
  mutate(status = ifelse(cell_type %in% filtered_cell_types, "Included", "Excluded")) %>%
  filter(!cell_type %in% exclude_celltypes)   # <- this line removes progenitors


# Plot with different colors
p <- plot_data %>%
  ggplot(aes(x = Fraction, fill = status)) +
  geom_histogram(bins = 30, color = "black") +
  facet_wrap(~ cell_type, scales = "fixed") +
  scale_fill_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Distribution of Cell Fractions Across Cell Types",
    x = "Fraction",
    y = "Count",
    fill = "Status"
  )

print(p)




p <- plot_data %>%
  ggplot(aes(x = Fraction, color = status, fill = status)) +
  geom_density(alpha = 0.3) +  # semi-transparent fill
  facet_wrap(~ cell_type, scales = "fixed") +
  scale_fill_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  scale_color_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Density of Cell Fractions Across Cell Types",
    x = "Fraction",
    y = "Density",
    fill = "Status",
    color = "Status"
  )
print(p)




library(dplyr)
library(ggplot2)
library(forcats)

# Reorder cell_type by mean Fraction
plot_data <- plot_data %>%
  group_by(cell_type) %>%
  mutate(mean_fraction = mean(Fraction)) %>%
  ungroup() %>%
  mutate(cell_type = fct_reorder(cell_type, mean_fraction))

# Plot
p <- plot_data %>%
  ggplot(aes(x = Fraction, y = cell_type, color = status)) +
  geom_point(position = position_jitter(height = 0.2, width = 0), alpha = 0.7) +
  scale_color_manual(values = c("Included" = "steelblue", "Excluded" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Cell Fractions Across Cell Types (Ordered by Mean Fraction)",
    x = "Fraction",
    y = "Cell Type",
    color = "Status"
  )

print(p)
```










#PCA
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Step 1: Filter out unwanted scores
analysis_data_filtered <- analysis_data 

# Step 2: Convert long to wide format (samples = rows, cell types = columns)
pca_data <- analysis_data_filtered %>%
  select(Sample, cell_type, Fraction) %>%
  pivot_wider(
    names_from = cell_type,
    values_from = Fraction,
    values_fill = 0  # fill missing fractions with 0
  ) %>%
  column_to_rownames("Sample")  # samples as rownames

# Step 3: Ensure all columns are numeric
pca_data[] <- lapply(pca_data, as.numeric)

# Step 4: Perform PCA
pca_result <- prcomp(pca_data, center = TRUE, scale. = TRUE)

# Step 5: Extract PCA scores
pca_df <- as.data.frame(pca_result$x)
pca_df$Sample <- rownames(pca_df)

# Optional: add metadata for coloring
metadata <- analysis_data %>%
  select(Sample, hpv_status) %>%
  distinct(Sample, .keep_all = TRUE)

pca_df <- pca_df %>%
  left_join(metadata, by = "Sample")

# Step 6: Plot PCA
ggplot(pca_df, aes(x = PC1, y = PC2, color = hpv_status)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "PCA of Filtered xCell Fractions",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100,1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100,1), "%)")
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")


```

#Create dataframe of explained variance
```{r}
explained_var <- summary(pca_result)$importance[2, ] * 100

explained_df <- data.frame(
  PC = names(explained_var),
  VarianceExplained = as.numeric(explained_var)
)

explained_df
```

#Which clinical variables is more associated with variability? Linear models with PC
```{r}
merged_df <- merge(pca_df, TP53status_clinical, by.x = "Sample", by.y = "id")

# We have two Hpv status, Keep the x version
merged_df$hpv_status <- merged_df$hpv_status.x

merged_df$p53_ihc_status <- relevel(factor(merged_df$p53_ihc_status), ref = "Wt")
merged_df$tumor_size_category <- relevel(factor(merged_df$tumor_size_category), ref = "≤25 mm")
merged_df$depth_invasion_category<- relevel(factor(merged_df$depth_invasion_category), ref = "≤5 mm")



# Load necessary packages
library(broom)

# Perform linear regression and tidy the results for each principal component (PC)
lm_results <- lapply(grep("^PC", names(merged_df), value = TRUE), function(pc) {
  # Fit the linear model with the specified clinical variables
  model <- lm(as.formula(paste(pc, "~ age + factor(figo_ini_ava) + factor(nodal_status) + factor(hpv_status) +factor(margins)+ factor(depth_invasion_category)+factor(tumor_size_category) + factor(p53_ihc_status)")), data = merged_df)
  
  # Tidy the model results
  tidy_model <- tidy(model)
  
  # Add the PC name to each tidy output for reference
  tidy_model$PC <- pc
  
  return(tidy_model)
})

# Combine all the results into one data frame
lm_results_df <- do.call(rbind, lm_results)

# View the results
head(lm_results_df)
```

```{r}
# Round numeric columns
lm_results_df_rounded <- lm_results_df %>%
  mutate(
    estimate   = round(estimate, 2),
    std.error  = round(std.error, 2),
    statistic  = round(statistic, 2),
    p.value    = round(p.value, 3)
  )

# Export to CSV
write.csv(lm_results_df_rounded, 
          file = "lm_results_HPV_xcell.csv", 
          row.names = FALSE)
```



#Variables contributing to PC1 and PC2?
```{r}
# Assuming pca_result contains the PCA object with loadings
# Extract the loadings for PC1 and PC2
loadings <- pca_result$rotation  # Loadings are in the rotation matrix

# Get the top 3 contributors to PC1 and PC2 (by absolute value)
top_PC1 <- head(sort(abs(loadings[, 1]), decreasing = TRUE), 3)
top_PC2 <- head(sort(abs(loadings[, 2]), decreasing = TRUE), 3)

# Combine these results for display and rename the first column as 'Variable'
top_contributors <- data.frame(
  Variable = c(names(top_PC1), names(top_PC2)),
  PC = c(rep("PC1", 3), rep("PC2", 3)),
  Loading = c(loadings[names(top_PC1), "PC1"], loadings[names(top_PC2), "PC2"])
)

top_contributors$Contribution_pct <- top_contributors$Loading^2 * 100


# Plot the top loadings for PC1 and PC2 as percentages
ggplot(top_contributors, aes(x = reorder(Variable, Contribution_pct), y = Contribution_pct, fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "grey27", "PC2" = "grey50")) +
  labs(title = "Top Contributions to PC1 and PC2 (%)", 
       x = "Variable", 
       y = "Contribution (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.margin = margin(10, 10, 10, 10),  # Adjust margins to reduce space
        axis.text = element_text(size = 8),  # Smaller axis text
        axis.title = element_text(size = 10), # Smaller axis titles
        panel.grid = element_blank())  # Remove grid for a cleaner look


```




#Final plot with top3 variables contributing to PC1 and PC2
```{r}
library(ggplot2)
library(ggrepel)

# Step 1: Extract PCA scores 
pca_scores <- pca_df  # Assuming PC1 and PC2 are already in here

# Step 2: Extract loadings (directions of variable contributions)
loadings <- pca_result$rotation[, 1:2]  # Only PC1 and PC2

# Step 3: Get top contributing variables to PC1 and PC2
top_vars <- unique(c(names(top_PC1), names(top_PC2)))  # Combine top contributors

# Scale loadings for visualization — use a constant or use your data's scale
loading_df <- as.data.frame(loadings[top_vars, ])
loading_df$Variable <- rownames(loading_df)


# Your pale colors for HPV status
hpv_colors_custom <- c(
  "Positive" = "#FF9999",  # pale red (HPV-associated)
  "Negative" = "#9999FF"   # pale blue (HPV-independent)
)


# Step 4: Create PCA plot with arrows (biplot-style)
ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(hpv_status))) +
  geom_point(size = 3) +
  # Add arrows for variable loadings
  geom_segment(data = loading_df,
               aes(x = 0, y = 0, xend = PC1 * 10, yend = PC2 * 10),  # scaled for visibility
               arrow = arrow(length = unit(0.25, "cm")), color = "black") +
  geom_text_repel(data = loading_df,
                  aes(x = PC1 * 10, y = PC2 * 10, label = Variable), 
                  color = "black", size = 3) +
  labs(title = "PCA with Top Variable Contributions",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2,2] * 100, 1), "%)"),
       color = "HPV Status") +
  scale_color_manual(values = hpv_colors_custom,
                     labels = c("Negative" = "HPV-independent", 
                                "Positive" = "HPV-associated")) +

  theme_minimal()
```



```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(hpv_status))) +
  geom_point(size = 4.0) +
  
  # Add arrows for variable loadings
  geom_segment(
    data = loading_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * 10,
      yend = PC2 * 10
    ),
    arrow = arrow(length = unit(0.25, "cm")),
    color = "black"
  ) +
  
  geom_text_repel(
    data = loading_df,
    aes(
      x = PC1 * 10,
      y = PC2 * 10,
      label = Variable
    ),
    color = "black",
    size = 3
  ) +
  
  labs(
    title = "PCA with Top Variable Contributions",
    x = paste0(
      "PC1 (",
      round(summary(pca_result)$importance[2,1] * 100, 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(summary(pca_result)$importance[2,2] * 100, 1),
      "%)"
    ),
    color = "HPV Status"
  ) +
  
  scale_color_manual(
    values = hpv_colors_custom,
    labels = c(
      "Negative" = "HPV-independent",
      "Positive" = "HPV-associated"
    )
  ) +
  
  theme_minimal()
```






#Calculate significant cell types only
```{r}
# Ensure HPV status is treated as a factor
analysis_data$hpv_status <- factor(
  analysis_data$hpv_status,
  levels = c("Positive", "Negative")
)

# Perform Mann-Whitney U / Wilcoxon rank-sum test for each cell type
results_hpv <- analysis_data %>%
  filter(!is.na(hpv_status), !is.na(Fraction)) %>%
  group_by(cell_type) %>%
  summarise(
    n_positive = sum(hpv_status == "Positive"),
    n_negative = sum(hpv_status == "Negative"),
    p_value = wilcox.test(Fraction ~ hpv_status)$p.value,
    median_positive = median(Fraction[hpv_status == "Positive"], na.rm = TRUE),
    median_negative = median(Fraction[hpv_status == "Negative"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Benjamini-Hochberg FDR correction across ALL tested cell types
    FDR = p.adjust(p_value, method = "BH")
  ) %>%
  arrange(FDR)

# View all results
results_hpv
```


```{r}
# Cell types significant after multiple-testing correction
significant_results <- results_hpv %>%
  filter(FDR < 0.05)

# Extract significant cell types
significant_cell_types <- significant_results$cell_type

# Merge significant results back with the original data
significant_data <- analysis_data %>%
  filter(cell_type %in% significant_cell_types) %>%
  left_join(
    significant_results %>%
      select(cell_type, p_value, FDR,
             median_positive, median_negative),
    by = "cell_type"
  )
```

```{r}
# ============================================================
# 1. Get FDR-significant cell types
# ============================================================

significant_results <- results_hpv %>%
  filter(FDR < 0.05) %>%
  arrange(FDR)

significant_cell_types <- significant_results$cell_type


# ============================================================
# 2. Prepare plotting data
# ============================================================

analysis_data_significant <- analysis_data %>%
  filter(
    cell_type %in% significant_cell_types,
    !is.na(hpv_status),
    !is.na(Fraction)
  ) %>%
  mutate(
    cell_type = factor(
      cell_type,
      levels = significant_cell_types
    ),
    hpv_status = factor(
      hpv_status,
      levels = c("Positive", "Negative")
    )
  )


# Check how many observations each cell type has
table(analysis_data_significant$cell_type)


# ============================================================
# 3. Prepare FDR labels
# ============================================================

plot_labels <- significant_results %>%
  mutate(
    cell_type = factor(
      cell_type,
      levels = significant_cell_types
    ),
    FDR_label = paste0(
      "FDR = ",
      format.pval(FDR, digits = 2, eps = 0.001)
    )
  )


# ============================================================
# 4. Plot
# ============================================================

HPVplot <- ggplot(
  analysis_data_significant,
  aes(
    x = hpv_status,
    y = Fraction,
    fill = hpv_status
  )
) +
  
geom_boxplot(
  width = 0.45,
  outlier.shape = NA
) +

geom_jitter(
  aes(color = hpv_status),
  width = 0.08,
  size = 1.5,
  alpha = 0.65
) +
  
facet_wrap(
  ~cell_type,
  scales = "free_y",
  nrow = 3,
  ncol = 4
)+
  
  geom_text(
    data = plot_labels,
    aes(
      x = 1.5,
      y = Inf,
      label = FDR_label
    ),
    inherit.aes = FALSE,
    vjust = 1.5,
    size = 3.5
  ) +
  
  scale_fill_manual(
    values = c(
      "Positive" = "#FF9999",
      "Negative" = "#9999FF"
    )
  ) +
  
  scale_color_manual(
    values = c(
      "Positive" = "#FF6666",
      "Negative" = "#6666CC"
    )
  ) +
  
  theme_minimal() +
  
  labs(
    title = "Cell Type Fractions by HPV Status",
    x = "HPV Status",
    y = "xCell Score",
    fill = "HPV Status",
    color = "HPV Status"
  ) +
  
  theme(
    panel.spacing = unit(0.8, "lines"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    strip.text = element_text(
      size = 11,
      face = "bold"
    ),
    plot.title = element_text(
      size = 16,
      face = "bold"
    )
  )

HPVplot
```


```{r}
# ============================================================
# HEATMAP: FDR-significant cell types only
# ============================================================

library(ComplexHeatmap)
library(tidyr)
library(dplyr)
library(tibble)

# ============================================================
# 1. Identify FDR-significant cell types
# ============================================================

significant_cell_types <- results_hpv %>%
  filter(FDR < 0.05) %>%
  arrange(FDR) %>%
  pull(cell_type)

# Check which cell types are significant
print(significant_cell_types)

# Number of significant cell types
length(significant_cell_types)
```

```{r}
# ============================================================
# 2. Keep only FDR-significant cell types
# ============================================================

heatmap_data <- analysis_data %>%
  filter(
    cell_type %in% significant_cell_types,
    !is.na(Fraction),
    !is.na(hpv_status)
  ) %>%
  select(
    Sample,
    Fraction,
    cell_type,
    hpv_status
  )

# Check number of samples per HPV group
table(
  heatmap_data$hpv_status
)

# Check number of observations per cell type
table(
  heatmap_data$cell_type
)
```

```{r}
# ============================================================
# 3. Create cell type × sample matrix
# ============================================================

data_grouped <- heatmap_data %>%
  group_by(cell_type, Sample) %>%
  summarise(
    Fraction = mean(Fraction, na.rm = TRUE),
    .groups = "drop"
  )

data_wide <- data_grouped %>%
  pivot_wider(
    names_from = Sample,
    values_from = Fraction
  ) %>%
  column_to_rownames("cell_type")

# Convert to matrix
data_matrix <- as.matrix(data_wide)

# Make sure values are numeric
mode(data_matrix) <- "numeric"
```




```{r}
# ============================================================
# 4. Z-score by cell type
# ============================================================

data_zscore <- t(
  apply(
    data_matrix,
    1,
    function(x) as.numeric(scale(x))
  )
)

# Preserve names
rownames(data_zscore) <- rownames(data_matrix)
colnames(data_zscore) <- colnames(data_matrix)
```


```{r}
# ============================================================
# 5. Create HPV status annotation
# ============================================================

sample_metadata <- heatmap_data %>%
  distinct(Sample, hpv_status) %>%
  filter(
    Sample %in% colnames(data_zscore)
  ) %>%
  mutate(
    hpv_status = recode(
      hpv_status,
      "Negative" = "HPV-independent",
      "Positive" = "HPV-associated"
    )
  )

# IMPORTANT:
# Reorder metadata to exactly match heatmap columns

sample_metadata <- sample_metadata %>%
  slice(
    match(
      colnames(data_zscore),
      Sample
    )
  )

# Check that the order is correct
stopifnot(
  sample_metadata$Sample == colnames(data_zscore)
)
```

```{r}
# ============================================================
# 6. HPV colors
# ============================================================

hpv_colors_custom <- c(
  "HPV-associated" = "#FF9999",
  "HPV-independent" = "#9999FF"
)

column_annotation <- HeatmapAnnotation(
  hpv_status = sample_metadata$hpv_status,
  col = list(
    hpv_status = hpv_colors_custom
  )
)
```

```{r}
# ============================================================
# 7. Final heatmap
# ============================================================

Heatmap(
  data_zscore,
  name = "Z-Score",
  
  row_names_side = "left",
  column_title = "Samples",
  row_title = "FDR-significant cell types",
  
  cluster_columns = TRUE,
  cluster_rows = TRUE,
  
  clustering_distance_rows = "manhattan",
  clustering_distance_columns = "manhattan",
  
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  
  top_annotation = column_annotation
)
```



#NOW for scores

```{r}
xcell <- deconvolute(counts_filtered_filtered, "xcell")
length(xcell$cell_type)

#We only obtain 39 cell types. This function automatically drops any cell type that has zero (or near-zero) signal across all samples, that's why we're seeing only 39 out of 64.

xcell_long <- xcell %>%
  pivot_longer(
    cols = -cell_type, # All columns except `cell_type`
    names_to = "Sample", 
    values_to = "Fraction"
  )

analysis_data <- xcell_long %>%
  left_join(TP53status_clinical, by = c( "Sample"="id"))
```

#without excluding scores:
```{r}
threshold <- 0.01      # fraction cutoff
min_presence <- 0.1   # 10% of samples must exceed threshold

 
# Cell types to manually exclude
#scores_exclude <- c("microenvironment score", "immune score", "stroma score")

# Keep original data
analysis_data_unfiltered <- analysis_data 
#  filter(!(cell_type %in% scores_exclude))


# Calculate proportion present
present_summary <- analysis_data_unfiltered %>%
  group_by(cell_type) %>%
  summarise(present = mean(Fraction > threshold), .groups = "drop")

# Filter cell types that meet criteria
filtered_cell_types <- present_summary %>%
  filter(present >= min_presence) %>%
  pull(cell_type)

# Get excluded cell types BEFORE filtering
excluded_cell_types <- setdiff(unique(analysis_data_unfiltered$cell_type),
                               filtered_cell_types)

print(excluded_cell_types)

# Keep only filtered types
analysis_data <- analysis_data_unfiltered %>%
  filter(cell_type %in% filtered_cell_types)

# Plot excluded ones if needed
excluded_data <- analysis_data_unfiltered %>%
  filter(cell_type %in% excluded_cell_types)
```


```{r}
# Cell types that are considered "scores"
scores_only <- c("microenvironment score", "immune score", "stroma score")

# Keep only score cell types
analysis_data_scores <- analysis_data %>%
  filter(cell_type %in% scores_only)

# Check the result
unique(analysis_data_scores$cell_type)

```




#Significant ones

```{r}
# ============================================================
# SCORE ANALYSIS: HPV-associated vs HPV-independent
# ============================================================

library(dplyr)
library(ggplot2)

# Keep only complete observations
scores_data <- analysis_data_scores %>%
  select(Sample, cell_type, Fraction, hpv_status) %>%
  filter(
    !is.na(hpv_status),
    !is.na(Fraction)
  ) %>%
  mutate(
    hpv_status = factor(
      hpv_status,
      levels = c("Positive", "Negative")
    )
  )

# Check the number of samples per HPV group
table(scores_data$hpv_status)

# Check that the three scores are present
unique(scores_data$cell_type)
```


```{r}
# ============================================================
# Wilcoxon tests + BH-FDR correction
# FDR is calculated across the 3 xCell scores
# ============================================================

results_scores <- scores_data %>%
  group_by(cell_type) %>%
  summarise(
    
    n_positive = sum(hpv_status == "Positive"),
    n_negative = sum(hpv_status == "Negative"),
    
    median_positive = median(
      Fraction[hpv_status == "Positive"],
      na.rm = TRUE
    ),
    
    median_negative = median(
      Fraction[hpv_status == "Negative"],
      na.rm = TRUE
    ),
    
    p_value = wilcox.test(
      Fraction ~ hpv_status
    )$p.value,
    
    .groups = "drop"
  ) %>%
  mutate(
    FDR = p.adjust(
      p_value,
      method = "BH"
    ),
    
    significant = FDR < 0.05
  ) %>%
  arrange(FDR)

# View results
results_scores
```




```{r}
# ============================================================
# Prepare plot labels
# ============================================================

pale_colors_fill <- c(
  "Positive" = "#FF9999",
  "Negative" = "#9999FF"
)

pale_colors_color <- c(
  "Positive" = "#FF6666",
  "Negative" = "#6666CC"
)

# Order scores
score_order <- c(
  "microenvironment score",
  "immune score",
  "stroma score"
)

scores_data <- scores_data %>%
  mutate(
    cell_type = factor(
      cell_type,
      levels = score_order
    )
  )

# Prepare FDR labels
score_labels <- results_scores %>%
  mutate(
    cell_type = factor(
      cell_type,
      levels = score_order
    ),
    FDR_label = paste0(
      "FDR = ",
      format.pval(
        FDR,
        digits = 2,
        eps = 0.001
      )
    )
  )
```


```{r}
# ============================================================
# FINAL SCORE PLOT
# ============================================================

HPVplot_scores_FDR <- ggplot(
  scores_data,
  aes(
    x = cell_type,
    y = Fraction,
    fill = hpv_status
  )
) +
  
  geom_boxplot(
    width = 0.45,
    outlier.shape = NA,
    position = position_dodge(width = 0.6)
  ) +
  
  geom_jitter(
    aes(color = hpv_status),
    size = 1.5,
    alpha = 0.65,
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.6
    )
  ) +
  
  # Add FDR values
  geom_text(
    data = score_labels,
    aes(
      x = cell_type,
      y = Inf,
      label = FDR_label
    ),
    inherit.aes = FALSE,
    vjust = 1.5,
    size = 3.5
  ) +
  
  scale_fill_manual(
    values = pale_colors_fill
  ) +
  
  scale_color_manual(
    values = pale_colors_color
  ) +
  
  scale_x_discrete(
    labels = c(
      "microenvironment score" = "Microenvironment score",
      "immune score" = "Immune score",
      "stroma score" = "Stroma score"
    )
  ) +
  
  theme_minimal() +
  
  labs(
    title = "xCell Scores by HPV Status",
    x = "Score Type",
    y = "xCell Score",
    fill = "HPV Status",
    color = "HPV Status"
  ) +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    plot.title = element_text(
      size = 16,
      face = "bold"
    )
  )

HPVplot_scores_FDR
```



